\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Code 2013},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
% Make links footnotes instead of hotlinks:
\renewcommand{\href}[2]{#2\footnote{\url{#1}}}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\title{Code 2013}

\begin{document}
\maketitle

\% Justin Le \% January 9, 2014

\emph{Originally posted on
\textbf{\href{https://blog.jle.im/entry/code-2013.html}{in Code}}.}

It's a bit late, but, inspired by my own admittedly egregious
\href{https://twitter.com/mstk/status/418179201091399680}{tweet} in
participation of the awesome \href{http://code2013.herokuapp.com/}{\#code2013}
hashtag, here is a review of 2013 in terms of the programming languages I've
worked on and a general wrap-up of the adventures that 2013 had to offer me. It
was quite surely the most productive/adventurous year of programming of my
entire life.

\section{Scientific Computing}\label{scientific-computing}

This year was the year I began to take scientific computing seriously. I had
already been using
\textbf{\href{http://www.wolfram.com/mathematica/}{Mathematica}} for many years,
for some basic modeling work on analytic and numeric systems. This year
continued my studies; however, I also enrolled in a two-quarter class taught by
the Physics department (my own field of study) at my university on scientific
computing techniques and applications in the context of Computational Physics.

However, this course would be relying heavily on parallel computing techniques
as well as low-level compiled languages like C and Fortran, with simulations
that would have decently long executing times.

\subsection{Linux}\label{linux}

I quickly realized that I could not do very much with my little Windows laptop,
so I procured for the first time in my life a persistent server. I used the
lowest tier from \href{http://www.togglebox.com/}{ToggleBox}, a great cloud
hosting service that gave me a year free. This was my return to unix-based
systems after a break of several years, and in doing this I sort of went crazy
again writing several
\textbf{\href{http://en.wikipedia.org/wiki/Bash_(Unix_shell)}{Bash}} scripts
that I am continuing to
\href{http://blog.jle.im/entry/log-sh-lightweight-command-line-note-logging}{open-source}/publish
today; many of my scripts revolved around the
\href{http://todotxt.com/}{todo.txt} framework.

I did some work on parsing some of my custom journal entry files and things like
that, and with this I returned to using
\textbf{\href{http://en.wikipedia.org/wiki/Awk}{awk}}, which I had worked with a
few years ago for processing logs. I rather really liked the language and it
seemed like a very perfectly adapted DSL. There are many problems that are
solved most easily in awk and I have a lot of respect for the language.

\subsection{Vim}\label{vim}

Also this year was the year I decided to begin taking
vi/\href{http://www.vim.org/}{vim} seriously. This is because I realized that I
would be doing most of my coding through an ssh connection, so I needed to pick
a good terminal editor. After a lot of research I settled on vim. I've been
exclusively using vim for some time now and, while it has a lot of endearing
flaws, I am happy with my choice. With a toolbelt of plugins and extensions as
well as a firm grasp of the vanilla basics, vim has served me very well. I have
even delved into some of my own personal plugin development in
\textbf{\href{http://en.wikipedia.org/wiki/Vim_script}{vimscript}}

\subsection{Computation}\label{computation}

For the aforementioned course, we had the opportunity to use either C or
Fortran. Because I had dabbled in C before and have worked in C-like languages
all my life, I figured I would take the opportunity to learn a new language. So
I learned \textbf{\href{http://en.wikipedia.org/wiki/Fortran}{Fortran}} ---
specifically, F77. This decision ended up being a good one because many of the
established numerical/computational libraries in Physics were written in the
seventies, and almost all of them are in Fortran. This interoperability was very
handy, and I was also able to dive into them and pick them apart.

Upon further digging into Fortran I found that I actually really liked the
semantics and syntax of the language. The language seemed to fit to the
computational techniques and algorithms I was learning about like a glove, more
so than with C or C++. Many things that felt awkward in C felt natural in
Fortran, and the entire program execution model just seemed seamless and
elegant.

I still consider Fortran among my favorite imperative languages today.

However, for group projects I was forced to work with
\textbf{\href{http://en.wikipedia.org/wiki/C++}{C++}}, specifically with its
\href{http://www.nvidia.com/object/cuda_home_new.html}{Cuda} hooks, on a
parallel N-body simulation. I came to appreciate the ad-hoc binary serialization
of data structures that many languages I had been using in the past lacked.

And here is the appropriate time to mention my work in
\textbf{\href{http://www.gnuplot.info/}{gnuplot}} for the presentation of the
data generated by these course projects

In another course on numerical analysis, there was much usage also of
\textbf{\href{http://www.mathworks.com/products/matlab/}{Matlab}}, if only to
demonstrate certain theorems and techniques. In addition to this, for some of my
lab courses, I did much of my data analysis and modeling/presentation in matlab.

\subsection{Java}\label{java}

I also took a beginning \textbf{\href{https://www.java.com/en/}{Java}} course
during my spring quarter. Mostly to fill in some requirements for my degree, I
didn't actually learn too much. This section is just to make this post complete
:) My most recent large-ish work on Java was for a Machine Learning/Artificial
Intelligence course the year before. No new insights from Java gleamed this year
that I can post here\ldots except that I can appreciate in a perverse way Java's
stubbornness to keep a consistent story of
\href{http://steve-yegge.blogspot.com/2006/03/execution-in-kingdom-of-nouns.html}{nouns}.

\section{Haskell}\label{haskell}

Around the end of winter I stumbled upon
\textbf{\href{http://www.haskell.org/haskellwiki/Haskell}{Haskell}}. I am not
completely sure how, actually. Perhaps it is just fate. But it is probably one
of the more significant life-changing discoveries of my programming life. This
event will define the rest of my entire year; it also inspired this blog itself.
Haskell is a beautifully expressive functional, pure, non-strict language that
merges serious and practical programming and powerful mathematics. Writing fully
about my reflections on Haskell is beyond the scope of this post. However, if
you don't yet know Haskell, why don't you give
\href{http://learnyouahaskell.com/}{Learn You A Haskell} a quick read? :) You
won't regret it!

I began with trying to implement my Fortran Monte Carlo programs in Haskell for
my course. I wasn't quite skilled enough to really use the full power of Haskell
before my computation course ended, I was able to really simplify my programs
and appreciate even a small part of the elegance of Haskell.

\subsection{Blog}\label{blog}

Over the summer I delved back into web development by deciding that my next big
Haskell project would be a blog engine. The result is this blog!

Actually, I have noticed that, more than for other languages, having and
maintaining a blog is a rite of passage and is sort or embedded into the culture
of the Haskell community. I also noticed that a lot of the work I was doing was
novel in the sense that I couldn't find any real established literature; I had
to forge a lot of the way myself. I wanted to be able to be able to give back to
the community (which is, by the way, one of the best programming language
communities I have found) and maybe jot down some of these struggles or
observations.

(This is probably also the best place to drop my usage of
\textbf{\href{http://www.graphviz.org/}{graphviz}}'s \emph{dot} DSL for some
diagrams for some blog posts. I like it a lot actually, and its declarative
model is pretty useful.)

(Oh yes I found myself learning and using some
\textbf{\href{http://en.wikipedia.org/wiki/SQL}{SQL}} for this if only to better
understand how to write joins in
\href{http://hackage.haskell.org/package/esqueleto}{Esqueleto})

\section{Web Development}\label{web-development}

I made a couple of forays into Web Development this year. This blog was one, but
there was very little front-end development work actually.

I made some small toy projects on
\textbf{\href{https://www.ruby-lang.org/en/}{ruby}} on my favorite framework,
\href{http://www.sinatrarb.com/}{Sinatra}. These are mostly just simple file
server and organization apps and GUI front-ends for native compiled programs
(simulations and the like). These all involved somewhat minimal front-end
\textbf{\href{http://en.wikipedia.org/wiki/JavaScript}{javascript}} and are
styled with \textbf{\href{http://sass-lang.com/}{sass}}, a CSS pre-processor
that I was at first hesitant to use but am not more or less completely dependent
on. (Maybe that was the reason I was hesitant in the first place? Oh well!)

I actually really enjoy ruby; before Haskell, it was my favorite language for
many years and my go-to scripting and one-off language. I also tended to
``think'' in ruby whenever approaching programming problems. I'll always have a
sort of nostalgic attachment to ruby, and in many cases you can't beat its
expressiveness and utility as a scripting language right off the bat (something
that it gains from its perl influences).

Some time in the fall, \href{http://www.intuit.com/}{Intuit} came to my
university and hosted a Hackathon. My team ended up rapidly prototyping our app
using ruby with the \href{http://rubyonrails.org/}{Rails} framework. So this is
my most recent web dev adventure. This was actaully my first time using Rails,
despite having used ruby for many years. The experience was sort of
disorienting, much magic. A lot felt very un-ruby. I guess I'll be sticking with
Sinatra for now :)

\section{Engineering}\label{engineering}

I had a summer job working at a Physics lab doing control theory work in the
field of electrical engineering. Much of it was in Matlab, especially with the
visual ``language'' of
\textbf{\href{http://www.mathworks.com/products/simulink/}{Simulink}}. I used
\textbf{\href{http://www.ni.com/labview/}{LabVIEW}} extensively as well; I guess
I didn't count those in my tweet but they probably deserve more recognition than
graphviz at least.

Simulink is a pretty amazing simulation system and I understand that it is very
effectively used in its industry fields. I was able to do a lot with its
platform and ecosystem right out of the box that would have taken much time to
shoddily replicate in another language. Its integration with Matlab (and
compiled Matlab) makes it extremely powerful for a very wide range of control
applications. Wow.

I also worked a lot on the \href{http://arduino.cc/}{Arduino} platform, using
\textbf{\href{http://en.wikipedia.org/wiki/C_(programming_language)}{C}}. It was
very nice to work with. This is before I really knew much about dynamic memory
allocation, but the work I did (simple image processing and pattern recognition
alongside startup procedures) did not require it.

\section{Fall Courses}\label{fall-courses}

\subsection{C/C++ Course}\label{cc-course}

I took a data structures in C and C++ course in the fall. It was pleasant
because while I had dabbled in both, I actually had very little formal
education. I was able to formalize my understanding of pointer-based data
structures (in C) and the syntax, language semantics, and language philosophy of
C++, being able to structure someone more-than-toy projects. I consider this my
first real exposure to formal C++ and manual dynamic memory management (without
the use of a garbage collector), and I gained a lot from slowing down to figure
this all out.

\subsection{Comparative Programming Languages
Course}\label{comparative-programming-languages-course}

And now the final chapter!

In my fall quarter I took my university's comparative programming languages
course.

\subsubsection{OCaml}\label{ocaml}

The first language we went into was \textbf{\href{http://ocaml.org/}{OCaml}}, a
member of the (typically) functional, strict, impure ML family of languages.
Learning it was a pleasant experience that helped me appreciate the advances in
language development that lead to Haskell. I was able to more formalized what I
liked and didn't like about Haskell. I grew an appreciation for Haskell's IO
model; I also thought it was interesting that OCaml did not have the stubborn
insistence on making type notation mirror data notation that Haskell does. The
Haskell tuple \texttt{(Int,\ Int)} is the OCaml tuple \texttt{int\ *\ int}; the
Haskell type application \texttt{Maybe\ Int} is in OCaml \texttt{int\ maybe}. It
gets rid of a lot of the confusion cased by mixing up type constructors and data
constructors in Haskell. But having the two be parallel does alow for some
interesting reasoning, and it also allows us to deal with Kinds in the same way
that we deal with Types.

But I'm getting a bit ahead of myself! And I am going on a tangent. There are a
lot of interesting differences between OCaml and Haskell :)

The culmination of our work in OCaml was using ocamlyacc to write an interpreter
of a simple subset of a functional language. It was a nice contrast to how in
Haskell we typically write our own parsers using parser combinators instead of
outsourcing to a ``compiled'' DSL.

\subsubsection{Python}\label{python}

Our next language was python! I do not have many good things to say about this
language :P I was always sort of a ruby fanboy. I have a long list of complaints
about python from a language design standpoint. But I'm here to praise its
benefits. The idiomatic usage of decorators was something interesting and very
python-esque. The object/class model was interesting in contrast to Javascript,
Ruby, Java/C\#, etc. There, I was nice :)

\subsubsection{Prolog}\label{prolog}

(\href{http://www.swi-prolog.org/}{SWI-})\textbf{\href{http://en.wikipedia.org/wiki/Prolog}{Prolog}}
was exciting to work with. I had a lot of fun with it. I'll just put
\href{http://c2.com/cgi/wiki?PrologLanguage}{this quote} here:

\begin{quote}
I used Prolog in a comparative languages course. The biggest program we did was
a map-coloring one (color a map with only four colors so that no bordering items
have the same color, given a mapping of things that border each other). I say
biggest because we were given the most time with it. I started out like most
people in my class trying to hack the language into letting me code a stinking
algorithm to color a stinking map. Then I wrote a test function to check if the
map was colored and, in a flash of prolog, realized that that was really all I
needed to code.
\end{quote}

Really though, the language was a nice joy. I talked to some people about bigger
applications and I read up on a way to run a web server with Prolog. It was a
little interesting but I felt like at some point the language model might start
interfering with its usages, instead of in Haskell where the language model
enhances much of its usage. Maybe that was just because I hadn't tried it
enough. Still, Prolog remains a great (domain specific?) language for solving
constraint problems and logic problems. Perhaps I will give it more of a look
some time in the future!

\section{2013 in Review}\label{in-review}

Welp, that's it! My observations on the languages I used this year, sort of
patched together to tell the story of my programming adventures this year.

What were some of your own adventures this year, in relation to new/old
languages?

\section{Signoff}\label{signoff}

Hi, thanks for reading! You can reach me via email at
\href{mailto:justin@jle.im}{\nolinkurl{justin@jle.im}}, or at twitter at
\href{https://twitter.com/mstk}{@mstk}! This post and all others are published
under the \href{https://creativecommons.org/licenses/by-nc-nd/3.0/}{CC-BY-NC-ND
3.0} license. Corrections and edits via pull request are welcome and encouraged
at \href{https://github.com/mstksg/inCode}{the source repository}.

If you feel inclined, or this post was particularly helpful for you, why not
consider \href{https://www.patreon.com/justinle/overview}{supporting me on
Patreon}, or a \href{bitcoin:3D7rmAYgbDnp4gp4rf22THsGt74fNucPDU}{BTC donation}?
:)

\end{document}
