\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{\textbf{#1}}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={The Const Applicative and Monoids},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
% Make links footnotes instead of hotlinks:
\renewcommand{\href}[2]{#2\footnote{\url{#1}}}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\title{The Const Applicative and Monoids}

\begin{document}
\maketitle

\% Justin Le \% May 8, 2018

\emph{Originally posted on
\textbf{\href{https://blog.jle.im/entry/const-applicative-and-monoids.html}{in
Code}}.}

The Applicative typeclass has a somewhat infamous reputation for having opaque
laws. There are a lot of great
\href{https://wiki.haskell.org/Typeclassopedia\#Alternative_formulation}{alternative}
\href{https://www.reddit.com/r/haskell/comments/2lompe/where_do_the_applicative_laws_come_from/clws90h/}{rephrasing}
of these laws, from many different approaches. For this post, however, I want to
talk about Applicative in terms of one of my favorites: \texttt{Const}.

\section{Const}\label{const}

The \texttt{Const} data type from the standard libraries is relatively simple as
far as newtypes go:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{newtype} \DataTypeTok{Const}\NormalTok{ w a }\OtherTok{=} \DataTypeTok{Const}\NormalTok{ \{}\OtherTok{ getConst ::}\NormalTok{ w \}}
\end{Highlighting}
\end{Shaded}

However, let's look at a less polymorphic version, \texttt{IntConst}, which is
essentially \texttt{Const\ Int}:\footnote{Note that if you want to play along in
  ghci, you should give this a \texttt{Show} instance by typing
  \texttt{deriving\ (Show)} after the data declaration}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{newtype} \DataTypeTok{IntConst}\NormalTok{ a }\OtherTok{=} \DataTypeTok{IntConst}\NormalTok{ \{}\OtherTok{ getIntConst ::} \DataTypeTok{Int}\NormalTok{ \}}
\end{Highlighting}
\end{Shaded}

For a \texttt{IntConst\ a}, the \texttt{a} is a \emph{phantom} type parameter.
This means that there are not necessarily any values of type \texttt{a} in a
value of type \texttt{IntConst\ a}. In modern GHC with PolyKinds, this means
that \texttt{a} might not even be a type that can have values --- you might
have, say, a value of type \texttt{IntConst\ Maybe}, or a value of type
\texttt{IntConst\ Monad}, and GHC would be perfectly happy.

\texttt{IntConst} admits a straightforward \texttt{Functor} instance that is a
lot like the \texttt{Functor} instance for \texttt{Proxy} and
\texttt{Either\ e}:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Functor} \DataTypeTok{IntConst} \KeywordTok{where}
    \FunctionTok{fmap}\NormalTok{ \_ (}\DataTypeTok{IntConst}\NormalTok{ w) }\OtherTok{=} \DataTypeTok{IntConst}\NormalTok{ w}
\end{Highlighting}
\end{Shaded}

In fact, sometimes I like to refer to \texttt{Const\ w\ a} as ``an
\texttt{Either\ w\ a} with only \texttt{Left}, no \texttt{Right}''. The
\texttt{Functor} instance reflects this pretty well:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Functor}\NormalTok{ (}\DataTypeTok{Either}\NormalTok{ e) }\KeywordTok{where}
    \FunctionTok{fmap}\NormalTok{ \_ (}\DataTypeTok{Left}\NormalTok{ e)  }\OtherTok{=} \DataTypeTok{Left}\NormalTok{ e        }\CommentTok{{-}{-} just like \textquotesingle{}Const\textquotesingle{}}
    \FunctionTok{fmap}\NormalTok{ f (}\DataTypeTok{Right}\NormalTok{ x) }\OtherTok{=} \DataTypeTok{Right}\NormalTok{ (f x)   }\CommentTok{{-}{-} who cares}
\end{Highlighting}
\end{Shaded}

However, the \texttt{Applicative} instance of \texttt{IntConst} is one of my
favorite things about it. Let's try to imagine how we'd write it.

First of all, let's look at the types of the functions we need:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{pure}\OtherTok{  ::}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ a}
\OtherTok{(\textless{}*\textgreater{}) ::} \DataTypeTok{IntConst}\NormalTok{ (a }\OtherTok{{-}\textgreater{}}\NormalTok{ b) }\OtherTok{{-}\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ b}
\end{Highlighting}
\end{Shaded}

Now, remember that \texttt{IntConst}'s type parameter is phantom, so we don't
have any actual values of type \texttt{a\ -\textgreater{}\ b}, \texttt{a}, or
\texttt{b} involved. An \texttt{IntConst\ a}, for any \texttt{a}, is really just
an \texttt{Int}. Essentially, once we strip out the newtype wrapper shenanigans
(replacing \texttt{IntConst\ a} with its contents, \texttt{Int}), we just get:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{pure}\OtherTok{  ::}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{Int}
\OtherTok{(\textless{}*\textgreater{}) ::} \DataTypeTok{Int} \OtherTok{{-}\textgreater{}} \DataTypeTok{Int} \OtherTok{{-}\textgreater{}} \DataTypeTok{Int}
\end{Highlighting}
\end{Shaded}

We now have a few options on how to implement these. Let's try one and see if it
works:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{IntConst} \KeywordTok{where}
    \FunctionTok{pure}\NormalTok{ \_                    }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{42}
    \DataTypeTok{IntConst}\NormalTok{ x }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ \_ }\OtherTok{=} \DataTypeTok{IntConst}\NormalTok{ x}
\end{Highlighting}
\end{Shaded}

A perfectly reasonable implementation, right? Our Applicative instance
type-checks. And if it type-checks, ship it! Time to call it a day and go home,
right?

Not quite.

\section{Applicative}\label{applicative}

Let's take a detour through the essense of the \texttt{Applicative} typeclass.
Or, at least, one particular interpretation of it that makes sense for instances
that represent some sort of effectful action.

The essence of Applicative, to me, is a way to combine values representing some
sort of ``effect'' in a sane way. In Haskell, we often talk about data types as
representing/describing some sort of effects. Applicative lets us combine
(``sequence'') them in a way that allows us to write powerful generic
combinators.

One way to look at it is as a generalization of \texttt{fmap} to taking two
parameters:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{fmap1 ::}\NormalTok{ (a }\OtherTok{{-}\textgreater{}}\NormalTok{ b     ) }\OtherTok{{-}\textgreater{}} \DataTypeTok{F}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{F}\NormalTok{ b}
\OtherTok{fmap2 ::}\NormalTok{ (a }\OtherTok{{-}\textgreater{}}\NormalTok{ b }\OtherTok{{-}\textgreater{}}\NormalTok{ c) }\OtherTok{{-}\textgreater{}} \DataTypeTok{F}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{F}\NormalTok{ b }\OtherTok{{-}\textgreater{}} \DataTypeTok{F}\NormalTok{ c}
\end{Highlighting}
\end{Shaded}

\texttt{fmap} alone lets you take a single \texttt{F\ a} and transform it into
an \texttt{F\ b}. \texttt{fmap2} (or, \texttt{liftA2} in the standard libraries)
is a way of taking \emph{two} \texttt{F}-values and squishing them into one fat
\texttt{F} value.

It does this by letting us talk about combining the \emph{effects} of an
\texttt{F\ a}, independent of the \texttt{a} (the result). For example,
\texttt{sequenceA\_}:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{sequenceA\_ ::} \DataTypeTok{Applicative}\NormalTok{ f }\OtherTok{=\textgreater{}}\NormalTok{ [f a] }\OtherTok{{-}\textgreater{}}\NormalTok{ f ()}
\end{Highlighting}
\end{Shaded}

Basically will take a list of \texttt{f\ a}s, and return a new \texttt{f\ ()}
that has \emph{all} of the effects of the \texttt{f\ a}s in the list.

To do this sensibly, we need also to talk about a ``no-op'' value:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{pure}\OtherTok{ ::}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{F}\NormalTok{ a}
\end{Highlighting}
\end{Shaded}

\texttt{pure\ x} is intended to be a no-op with ``no effects''.

With this, we can say something about the behavior of
\texttt{\textless{}*\textgreater{}} or \texttt{fmap2} or \texttt{liftA2}.
Namely:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The effects of \texttt{f\ \textless{}*\textgreater{}\ x} must have the effects
  of \texttt{f} once \emph{and} the effects of \texttt{x} once -- no more, and
  no less.
\item
  \texttt{pure}'s results must have no ``effects'', and so if used with
  \texttt{\textless{}*\textgreater{}}, introduces no extra effects:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{pure}\NormalTok{ f }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ x }\OtherTok{=} \FunctionTok{fmap}\NormalTok{ f x}
\NormalTok{f }\OperatorTok{\textless{}*\textgreater{}} \FunctionTok{pure}\NormalTok{ x }\OtherTok{=} \FunctionTok{fmap}\NormalTok{ (}\OperatorTok{$}\NormalTok{ x) f}
\end{Highlighting}
\end{Shaded}

  (Remember that \texttt{fmap} is not allowed to affect ``effects'' in any way)
\item
  Combining effects must be associative.
\end{enumerate}

With this guarantee, we can write \texttt{sequenceA\_} in a polymorphic way:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{sequenceA\_ ::} \DataTypeTok{Applicative}\NormalTok{ f }\OtherTok{=\textgreater{}}\NormalTok{ [f a] }\OtherTok{{-}\textgreater{}}\NormalTok{ f ()}
\NormalTok{sequenceA\_ []     }\OtherTok{=} \FunctionTok{pure}\NormalTok{ ()}
\NormalTok{sequenceA\_ (x}\OperatorTok{:}\NormalTok{xs) }\OtherTok{=}\NormalTok{ x }\OperatorTok{*\textgreater{}}\NormalTok{ sequenceA\_ xs}
\end{Highlighting}
\end{Shaded}

And we can say with certainty that:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Each ``effect'' of each value in the \texttt{{[}f\ a{]}} will be executed
  \emph{exactly} once: no more, and no less.
\item
  \texttt{sequenceA\_} of an empty list has no effects.
\end{enumerate}

This makes \texttt{sequenceA\_} a \emph{useful} combinator. The fact that we can
talk about how \texttt{sequenceA\_} behaves for all Applicative instances makes
it something that is \emph{worth} defining. If you use \texttt{sequenceA\_} for
your type, you can do it knowing that it will behave in a well-defined way: it
\emph{must} execute every action once (no more, and no less), and sequencing an
empty list \emph{must} have no effects.

If it weren't for those Applicative laws and expectations, \texttt{sequenceA\_}
would be a pretty useless function. It might completely ignore all effects, or
it might perform some of the effects multiple times\ldots who knows! The fact
that we have Applicative laws and expectations means we can look at the
implementation of \texttt{sequenceA\_} and know with certainty (and make bold
claims about) how \texttt{sequenceA\_} combines effects.

For example, we can always make the program substitution:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sequenceA\_ xs }\OperatorTok{*\textgreater{}}\NormalTok{ sequenceA\_ ys}
\CommentTok{{-}{-} can be substituted with}
\NormalTok{sequenceA\_ (xs }\OperatorTok{++}\NormalTok{ ys)}
\end{Highlighting}
\end{Shaded}

If \texttt{sequenceA\_} combines all the effects in \texttt{xs} once, and
\texttt{sequenceA\_} combines all the effects in \texttt{ys} once, and
\texttt{*\textgreater{}} combines the effects of either side once, then this is
a \emph{legal substitution} that doesn't change what your program does.

However, if the \texttt{Applicative} instance doesn't have any rules, we can't
do this. For example, the original form uses \texttt{pure\ ()} \emph{twice}
(once for each list's \texttt{{[}{]}} end), and the second form uses
\texttt{pure\ ()} \emph{once} (since there's only one list). If
\texttt{pure\ ()} was allowed to have effects\ldots then the first version would
have more effects than the second.

\section{Back to Const}\label{back-to-const}

With this new information in mind, let's revisit our instance for
\texttt{IntConst}.

In order for \texttt{IntConst}'s Applicative instance to behave meaningfully,
and in order to be able to match with user expectations of \texttt{Applicative}
instance, you need to make sure it follows the basic principles I mentioned
earlier (the effects of \texttt{f\ \textless{}*\textgreater{}\ x} has the
effects of \texttt{f} and \texttt{x} exactly once, and the properties about
\texttt{pure}).

We haven't defined what the ``effects'' of \texttt{IntConst} are yet, but let's
at least look at if our \texttt{pure} behaves sensibly with
\texttt{\textless{}*\textgreater{}}. Namely, let's check
\texttt{pure\ f\ \textless{}*\textgreater{}\ x\ =\ fmap\ f\ x}.

Note that this is a meaningful starting point because \texttt{fmap}'s definition
is \emph{fixed}. For any type, there is \emph{at most one} possible
\texttt{fmap} that is legal and lawful --- and in Haskell, we only have to check
that \texttt{fmap\ id} leaves all inputs unchanged.\footnote{There are other
  laws, but because of parametric polymorphism in Haskell, we know they must be
  true if and only if \texttt{fmap\ id\ =\ id}.}

With that out of the way, let's check our
\texttt{pure\ f\ \textless{}*\textgreater{}\ x\ =\ fmap\ f\ x} law with a simple
example for \texttt{x}\ldots say, \texttt{IntConst\ 5}. On the left hand side,
we have:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} pure \_                     = IntConst 42}
\CommentTok{{-}{-} IntConst x \textless{}*\textgreater{} InstConst \_ = IntConst x}
\FunctionTok{pure}\NormalTok{ f }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{5} \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{42} \OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{5}
                      \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{42}
\end{Highlighting}
\end{Shaded}

On the right hand side, we have:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} fmap f (IntConst x) = IntConst x}
\FunctionTok{fmap}\NormalTok{ f (}\DataTypeTok{IntConst} \DecValTok{5}\NormalTok{) }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{5}
\end{Highlighting}
\end{Shaded}

It is clear that this definition does not work, since \texttt{IntConst\ 42} is
not the same as \texttt{IntConst\ 5}.

What if we defined:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{IntConst} \KeywordTok{where}
    \FunctionTok{pure}\NormalTok{ \_                    }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{42}
    \DataTypeTok{IntConst}\NormalTok{ \_ }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ y }\OtherTok{=} \DataTypeTok{IntConst}\NormalTok{ y}
\end{Highlighting}
\end{Shaded}

Is that any better? Well,
\texttt{pure\ f\ \textless{}*\textgreater{}\ IntConst\ 5} is now equal to
\texttt{IntConst\ 5}, so that works out. But what about
\texttt{f\ \textless{}*\textgreater{}\ pure\ x\ =\ fmap\ (\$\ x)\ f}? Let's use
\texttt{IntConst\ 3} as our \texttt{f}. On the left hand side:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} pure \_                    = IntConst 42}
\CommentTok{{-}{-} IntConst \_ \textless{}*\textgreater{} IntConst y = IntConst y}
\DataTypeTok{IntConst} \DecValTok{3} \OperatorTok{\textless{}*\textgreater{}} \FunctionTok{pure}\NormalTok{ x }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{3} \OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{42}
                      \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{42}
\end{Highlighting}
\end{Shaded}

And on the right hand side:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} fmap f (IntConst x) = IntConst x}
\FunctionTok{fmap}\NormalTok{ (}\OperatorTok{$}\NormalTok{ x) (}\DataTypeTok{IntConst} \DecValTok{3}\NormalTok{) }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{3}
\end{Highlighting}
\end{Shaded}

Ah, that's wrong too, then.

At this point it might seem like I am facetiously moving very slowly to an
answer that has to use \emph{both} inputs. After all, my earlier statement
claimed that \texttt{f\ \textless{}*\textgreater{}\ x} has to use both the
effects of \texttt{f} and the effects of \texttt{x}, each exactly once. Because
we didn't really know what the ``effects'' of \texttt{IntConst} are, we don't
know exactly ``how'' to combine them\ldots but we can probably guess it has to
use both \texttt{Int}s. So, with that in mind, let's try another definition:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{IntConst} \KeywordTok{where}
    \FunctionTok{pure}\NormalTok{ \_                    }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{42}
    \DataTypeTok{IntConst}\NormalTok{ x }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ y }\OtherTok{=} \DataTypeTok{IntConst}\NormalTok{ (x }\OperatorTok{+}\NormalTok{ y)}
\end{Highlighting}
\end{Shaded}

Alright, now we use both \texttt{x} and \texttt{y} in the result. Let's see
again if this follows our expectations about \texttt{pure} -- if
\texttt{pure\ f\ \textless{}*\textgreater{}\ x} is the same as
\texttt{fmap\ f\ x}. Using \texttt{IntConst\ 5} again as \texttt{x}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} pure \_                     = IntConst 42}
\CommentTok{{-}{-} IntConst x \textless{}*\textgreater{} InstConst y = IntConst (x + y)}
\FunctionTok{pure}\NormalTok{ f }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{5} \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{42} \OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{5}
                      \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{47}
\end{Highlighting}
\end{Shaded}

On the right hand side, we have:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} fmap f (IntConst x) = IntConst x}
\FunctionTok{fmap}\NormalTok{ f (}\DataTypeTok{IntConst} \DecValTok{5}\NormalTok{) }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{5}
\end{Highlighting}
\end{Shaded}

Another dead end. It looks like it isn't just enough that we \emph{use} both
\texttt{Int}s\ldots we have to use them in a way such that the \texttt{Int} we
use as the result of \texttt{pure\ f} as to be an \emph{identity} to our
operation. Whatever \texttt{Int} is returned by \texttt{pure\ f} has to leave
any other \texttt{Int} unchanged when used with
\texttt{\textless{}*\textgreater{}}.

Thinking back, we remember that if our operation is \texttt{+}, we can use
\texttt{0}, since \texttt{0\ +\ x\ =\ x} and \texttt{x\ +\ 0\ =\ x}, for all
\texttt{x}. Luckily, our operation \texttt{x\ +\ y} is one that even \emph{has}
an identity. If we had chosen another operation (like \texttt{x\ +\ 2\ *\ y}),
we wouldn't be so lucky. Finally:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{IntConst} \KeywordTok{where}
    \FunctionTok{pure}\NormalTok{ \_                    }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{0}
    \DataTypeTok{IntConst}\NormalTok{ x }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ y }\OtherTok{=} \DataTypeTok{IntConst}\NormalTok{ (x }\OperatorTok{+}\NormalTok{ y)}
\end{Highlighting}
\end{Shaded}

At last this feels like something that should make sense. And, does it? Testing
out, again, \texttt{pure\ f\ \textless{}*\textgreater{}\ x\ =\ fmap\ f\ x}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} pure \_                     = IntConst 0}
\CommentTok{{-}{-} IntConst x \textless{}*\textgreater{} InstConst y = IntConst (x + y)}
\FunctionTok{pure}\NormalTok{ f }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{5} \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{0} \OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{5}
                      \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{5}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} fmap f (IntConst x) = IntConst x}
\FunctionTok{fmap}\NormalTok{ f (}\DataTypeTok{IntConst} \DecValTok{5}\NormalTok{) }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{5}
\end{Highlighting}
\end{Shaded}

Perfect! And, checking now
\texttt{f\ \textless{}*\textgreater{}\ pure\ x\ =\ fmap\ (\$\ x)\ f}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} pure \_                    = IntConst 0}
\CommentTok{{-}{-} IntConst x \textless{}*\textgreater{} IntConst y = IntConst (x + y)}
\DataTypeTok{IntConst} \DecValTok{3} \OperatorTok{\textless{}*\textgreater{}} \FunctionTok{pure}\NormalTok{ x }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{3} \OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{0}
                      \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{3}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} fmap f (IntConst x) = IntConst x}
\FunctionTok{fmap}\NormalTok{ (}\OperatorTok{$}\NormalTok{ x) (}\DataTypeTok{IntConst} \DecValTok{3}\NormalTok{) }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{3}
\end{Highlighting}
\end{Shaded}

This definition works for both\footnote{Note that in the real world we also have
  to verify that our definition combines effects in an \emph{associative} way,
  but we won't go too deeply into this for this article.}!

\subsection{The Effect of Const}\label{the-effect-of-const}

With our definition picked out, what do we think \texttt{sequenceA\_} does for
\texttt{IntConst}?

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{sequenceA\_ ::}\NormalTok{ [}\DataTypeTok{IntConst}\NormalTok{ a] }\OtherTok{{-}\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ ()}
\end{Highlighting}
\end{Shaded}

Well, if each application of \texttt{\textless{}*\textgreater{}} adds together
the \texttt{Int} in the \texttt{IntConst\ a}, and \texttt{sequenceA\_} uses
\texttt{\textless{}*\textgreater{}} once per every \texttt{IntConst\ a}\ldots we
can guess that \texttt{sequenceA\_} for \texttt{IntConst} is just \texttt{sum}!

This might be more clear if we strip away the newtype wrappers (replacing
\texttt{IntConst\ a} with its contents, \texttt{Int}):

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{sequenceA\_ ::}\NormalTok{ [}\DataTypeTok{Int}\NormalTok{] }\OtherTok{{-}\textgreater{}} \DataTypeTok{Int}
\end{Highlighting}
\end{Shaded}

From this definition of \texttt{\textless{}*\textgreater{}}, we can form an idea
of what the effects of the \texttt{IntConst} Applicative are: they \emph{add} to
some accumulator environment! And \texttt{pure\ \_\ =\ IntConst\ 0} means that
\texttt{pure\ \_} adds zero to our accumulator -- it leaves our accumulator
\emph{unchanged}, and so effectively has no effect.

That's why \texttt{sequenceA\_} is \texttt{sum} -- it sequences every effect of
the \texttt{IntConst}, which means that it sequences all of those ``add to the
accumulator'' effects one after the other.

\subsection{Alternative Pictures}\label{alternative-pictures}

Note that the requirements we gave for the \texttt{Applicative} instance doesn't
necessarily imply that the one we have is the only instance. For example, the
following instance is also valid:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{IntConst} \KeywordTok{where}
    \FunctionTok{pure}\NormalTok{ \_                    }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{1}
    \DataTypeTok{IntConst}\NormalTok{ x }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst}\NormalTok{ y }\OtherTok{=} \DataTypeTok{IntConst}\NormalTok{ (x }\OperatorTok{*}\NormalTok{ y)}
\end{Highlighting}
\end{Shaded}

If our ``combining'' action is \texttt{*}, then \texttt{pure} has to be an
identity. So, \texttt{pure\ \_\ =\ IntConst\ 1} works fine as an identity here,
since \texttt{1\ *\ x\ =\ x} and \texttt{x\ *\ 1\ =\ x}, for all \texttt{x}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} pure \_                     = IntConst 1}
\CommentTok{{-}{-} IntConst x \textless{}*\textgreater{} InstConst y = IntConst (x * y)}
\FunctionTok{pure}\NormalTok{ f }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{5} \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{1} \OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{IntConst} \DecValTok{5}
                      \OtherTok{=} \DataTypeTok{IntConst} \DecValTok{5}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} fmap f (IntConst x) = IntConst x}
\FunctionTok{fmap}\NormalTok{ f (}\DataTypeTok{IntConst} \DecValTok{5}\NormalTok{) }\OtherTok{=} \DataTypeTok{IntConst} \DecValTok{5}
\end{Highlighting}
\end{Shaded}

\subsection{A General Alternative}\label{a-general-alternative}

It looks like the Applicative for \texttt{IntConst} has to follow some pattern:

\begin{itemize}
\tightlist
\item
  \texttt{\textless{}*\textgreater{}} has to combine the \texttt{Int}s inside
  somehow using some operation \texttt{f}. (\texttt{f} also has to be
  associative, which is a point we didn't touch on specifically)
\item
  The \texttt{Int} that \texttt{pure} returns has to be an \emph{identity} to
  \texttt{f}.
\end{itemize}

Sound familiar?

This is all satisfied if and only if \texttt{f} and the result of \texttt{pure}
form a
\textbf{\href{https://www.schoolofhaskell.com/user/mgsloan/monoids-tour}{monoid}}
on the integers!

There is a very fundamental link here: the \texttt{Applicative} laws for
\texttt{IntConst} are satisfied if and only if
\texttt{\textless{}*\textgreater{}} acts monoidally on the contents, with
\texttt{pure}'s result as the identity of that monoid.

(For those unfamiliar with monoids, a \texttt{Monoid} in Haskell is a type
\texttt{w} that has an associative operation
\texttt{(\textless{}\textgreater{})\ ::\ w\ -\textgreater{}\ w\ -\textgreater{}\ w}
along with an identity \texttt{mempty\ ::\ w} that leaves values unchanged when
used with \texttt{\textless{}\textgreater{}}.)

So, \emph{any} \texttt{f} and \texttt{pure} works, as long as they form a
\emph{monoid}. And any monoid in the Integers is a valid \texttt{Applicative}
instance for \texttt{IntConst}!

\section{General Const}\label{general-const}

Let's revisit our original \texttt{Const} type:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{newtype} \DataTypeTok{Const}\NormalTok{ w a }\OtherTok{=} \DataTypeTok{Const}\NormalTok{ \{}\OtherTok{ getConst ::}\NormalTok{ w \}}
\end{Highlighting}
\end{Shaded}

The \texttt{Functor} instance is unique, so there isn't any leeway we have
(\texttt{fmap} is always fixed for every type):

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Functor}\NormalTok{ (}\DataTypeTok{Const}\NormalTok{ w) }\KeywordTok{where}
    \FunctionTok{fmap}\NormalTok{ \_ (}\DataTypeTok{Const}\NormalTok{ w) }\OtherTok{=} \DataTypeTok{Const}\NormalTok{ w}
\end{Highlighting}
\end{Shaded}

This is the only definition that preserves \texttt{fmap\ id\ =\ id}.

Now we can actually write an \texttt{Applicative} instance for
\texttt{Const\ w}\ldots as long as we provide a \texttt{Monoid} to use with
\texttt{w}\footnote{Note that the Applicative laws are loose enough to allow a
  different definition, with the same \texttt{pure}, but with
  \texttt{Const\ x\ \textless{}*\textgreater{}\ Const\ y\ =\ Const\ (y\ \textless{}\textgreater{}\ x}).
  But, this is just a different \texttt{Monoid} (\texttt{Const\ (Dual\ w)}).}!

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Monoid}\NormalTok{ w }\OtherTok{=\textgreater{}} \DataTypeTok{Applicative}\NormalTok{ (}\DataTypeTok{Const}\NormalTok{ w) }\KeywordTok{where}
    \FunctionTok{pure}\NormalTok{ \_              }\OtherTok{=} \DataTypeTok{Const} \FunctionTok{mempty}
    \DataTypeTok{Const}\NormalTok{ x }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{Const}\NormalTok{ y }\OtherTok{=} \DataTypeTok{Const}\NormalTok{ (x }\OperatorTok{\textless{}\textgreater{}}\NormalTok{ y)}
\end{Highlighting}
\end{Shaded}

Like how we said, as long as the ``combining'' function for \texttt{x} and
\texttt{y} have the identity that is given by the result of \texttt{pure}, this
is a valid Applicative.

The ``effects'' of this Applicative instance are ``accumulate to some
accumulator''. If this sounds familiar, this is because this is exactly the
effect of the \texttt{Writer\ w} Applicative instance. \texttt{Const\ w} and
\texttt{Writer\ w} have the same effects (``accumulate to some accumulator''),
and this can be seen clearly by comparing the two types:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Const}\NormalTok{  w a }\OtherTok{=} \DataTypeTok{Const}\NormalTok{  w}
\KeywordTok{data} \DataTypeTok{Writer}\NormalTok{ w a }\OtherTok{=} \DataTypeTok{Writer}\NormalTok{ w a}
\end{Highlighting}
\end{Shaded}

(\texttt{Const} is just \texttt{Writer} without the \texttt{a} value)

\section{What is Applicative Really?}\label{what-is-applicative-really}

If you think about this, this seems like a bit of a crazy coincidence.
Applicatives are an interesting concept, and Monoids are a different one.

But it looks like in order to make an \texttt{Applicative} instance for
\texttt{Const\ w}, the behavior of \texttt{\textless{}*\textgreater{}} and
\texttt{pure} have to follow some certain properties in order to fit the
Applicative laws\ldots and those properties are \emph{exactly} monoidal
properties and the Monoid laws.

To illustrate this link, we can see the type of \texttt{pure} and
\texttt{\textless{}*\textgreater{}} for \texttt{Const\ w}, without the newtype
wrappers (and ignored arguments)

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Monoid}\NormalTok{ w }\OtherTok{=\textgreater{}} \DataTypeTok{Applicative}\NormalTok{ (}\DataTypeTok{Const}\NormalTok{ w) }\KeywordTok{where}
\OtherTok{    pure  ::}\NormalTok{ w}
\OtherTok{    (\textless{}*\textgreater{}) ::}\NormalTok{ w }\OtherTok{{-}\textgreater{}}\NormalTok{ w }\OtherTok{{-}\textgreater{}}\NormalTok{ w}
\end{Highlighting}
\end{Shaded}

And let's look at the \texttt{Monoid} typeclass:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Monoid}\NormalTok{ w }\KeywordTok{where}
\OtherTok{    mempty ::}\NormalTok{ w}
\OtherTok{    (\textless{}\textgreater{})   ::}\NormalTok{ w }\OtherTok{{-}\textgreater{}}\NormalTok{ w }\OtherTok{{-}\textgreater{}}\NormalTok{ w}
\end{Highlighting}
\end{Shaded}

It seems like \texttt{Const} is nothing more than a (type-level) function on a
Monoid. As an \texttt{*\ -\textgreater{}\ (k\ -\textgreater{}\ *)}, it takes a
\texttt{*}-kinded Monoid and turns it into a
\texttt{k\ -\textgreater{}\ *}-kinded Monoid:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Monoid}\NormalTok{ (}\OtherTok{w ::} \OperatorTok{*}\NormalTok{) }\OtherTok{=\textgreater{}} \DataTypeTok{Applicative}\NormalTok{ (}\DataTypeTok{Const}\OtherTok{ w ::}\NormalTok{ k }\OtherTok{{-}\textgreater{}} \OperatorTok{*}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

``Give me a \texttt{Monoid} and I'll give you something
\texttt{k\ -\textgreater{}\ *} that is also a monoid!''

\texttt{Const} is a
\emph{\href{https://bartoszmilewski.com/2015/07/21/free-monoids/}{monoid
homomorphism}}: it takes a monoid \texttt{w} with \texttt{mempty} and
\texttt{(\textless{}\textgreater{})}, and turns it into a monoid
\texttt{Const\ w} with \texttt{pure\ \_} and
\texttt{\textless{}*\textgreater{}}:

\begin{Shaded}
\begin{Highlighting}[]
\DataTypeTok{Const}\NormalTok{ (x }\OperatorTok{\textless{}\textgreater{}}\NormalTok{ y) }\OtherTok{=} \DataTypeTok{Const}\NormalTok{ x }\OperatorTok{\textless{}*\textgreater{}} \DataTypeTok{Const}\NormalTok{ y}
\DataTypeTok{Const} \FunctionTok{mempty}   \OtherTok{=} \FunctionTok{pure}\NormalTok{ ()}
\end{Highlighting}
\end{Shaded}

Meaning ``\texttt{\textless{}\textgreater{}} then \texttt{Const}'' is the same
as ``\texttt{Const} then \texttt{\textless{}*\textgreater{}}'', and
``\texttt{Const} the \texttt{mempty}'' is the same as \texttt{pure\ ()}. Both
things essentially convey the exact same monoid -- one with
\texttt{\textless{}\textgreater{}} and \texttt{mempty}, and the other with
\texttt{\textless{}*\textgreater{}} and \texttt{pure\ ()}. In fact, it's a bit
more than a monoid homomorphism -- it's a \textbf{monoid isomorphism}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{getConst x }\OperatorTok{\textless{}\textgreater{}}\NormalTok{ getConst y }\OtherTok{=}\NormalTok{ getConst (x }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ y)}
\FunctionTok{mempty}                   \OtherTok{=}\NormalTok{ getConst (}\FunctionTok{pure}\NormalTok{ ())}
\end{Highlighting}
\end{Shaded}

Which means ``\texttt{getConst} then \texttt{\textless{}\textgreater{}}'' is the
same as ``\texttt{\textless{}*\textgreater{}} then \texttt{getConst}'', and
\texttt{mempty} is the same as \texttt{getConst\ (pure\ ())}. \texttt{getConst}
takes you from one monoid (with \texttt{\textless{}*\textgreater{}} and
\texttt{pure\ ()}) to another (with \texttt{\textless{}\textgreater{}} and
\texttt{mempty}).

One incidental observation -- \texttt{sequenceA\_} for \texttt{Const\ w} might
look familiar:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{sequenceA\_ ::} \DataTypeTok{Monoid}\NormalTok{ w }\OtherTok{=\textgreater{}}\NormalTok{ [}\DataTypeTok{Const}\NormalTok{ w a] }\OtherTok{{-}\textgreater{}} \DataTypeTok{Const}\NormalTok{ w ()}

\CommentTok{{-}{-} strip out newtype wrappers}
\OtherTok{sequenceA\_ ::} \DataTypeTok{Monoid}\NormalTok{ w }\OtherTok{=\textgreater{}}\NormalTok{ [w] }\OtherTok{{-}\textgreater{}}\NormalTok{ w}
\end{Highlighting}
\end{Shaded}

It's just \texttt{mconcat}!

As an exercise, see if you can understand this definition of \texttt{mconcat} in
terms of \texttt{Const} and \texttt{traverse}:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{mconcat}\OtherTok{ ::} \DataTypeTok{Monoid}\NormalTok{ w }\OtherTok{=\textgreater{}}\NormalTok{ [w] }\OtherTok{{-}\textgreater{}}\NormalTok{ w}
\FunctionTok{mconcat} \OtherTok{=}\NormalTok{ getConst }\OperatorTok{.}\NormalTok{ traverse\_ }\DataTypeTok{Const}
\end{Highlighting}
\end{Shaded}

\texttt{traverse\_}, if you aren't familiar with it, an ``effectful'' function
(in our case, \texttt{Const\ ::\ w\ -\textgreater{}\ Const\ w\ w}) over all
values in a container, and sequences all of their effects.

\subsection{Monoid is the Key}\label{monoid-is-the-key}

All of this actually witnesses the core of Applicative. A lot of people describe
Applicative as a ``lax monoidal functor''.

In this post, I was really handwavey with how I talked about ``effects''
(``\texttt{f\ \textless{}*\textgreater{}\ x} must use the effects of \texttt{f}
and \texttt{x} each once and only once'', I claimed, without defining what an
effect was). The notion of what an ``effect'' is really comes from each
individual Applicative, and each type really has its own conceptual picture of
what counts as an effect. The rigorous test of what is a meaningful way to have
an effect that can be combined comes from those laws
(\texttt{pure\ f\ \textless{}*\textgreater{}\ x\ =\ fmap\ f\ x}, etc.) and the
overall sentiment that the combination of effects is \emph{monoidal}.

At its heart, Applicative enforces that \texttt{liftA2} and
\texttt{\textless{}*\textgreater{}} are supposed to be ``monoidal'' in some way.
This fact is hidden by the normal form of the Applicative laws, but I feel like
seeing this play out in the \texttt{Applicative} instance for \texttt{Const} ---
how \texttt{Monoid} is exactly the constraint necessary to implement the
instance, and how \texttt{Const} forms a monoid isomorphism --- really helps
hammer in the monoidal nature of \emph{all} Applicative instances.

Applicative instances must be monoidal in how they sequence their effects.
Because \texttt{Const}'s effects are so simple (``accumulate a value''), this
makes it an especially obvious demonstration of this.

Hopefully this helps you gain some sense of appreciation between the link
between \texttt{Applicative} and \texttt{Monoid}, and also why \texttt{Const}'s
Applicative instance is defined the way it is!

\section{Signoff}\label{signoff}

Hi, thanks for reading! You can reach me via email at
\href{mailto:justin@jle.im}{\nolinkurl{justin@jle.im}}, or at twitter at
\href{https://twitter.com/mstk}{@mstk}! This post and all others are published
under the \href{https://creativecommons.org/licenses/by-nc-nd/3.0/}{CC-BY-NC-ND
3.0} license. Corrections and edits via pull request are welcome and encouraged
at \href{https://github.com/mstksg/inCode}{the source repository}.

If you feel inclined, or this post was particularly helpful for you, why not
consider \href{https://www.patreon.com/justinle/overview}{supporting me on
Patreon}, or a \href{bitcoin:3D7rmAYgbDnp4gp4rf22THsGt74fNucPDU}{BTC donation}?
:)

\end{document}
