\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{\textbf{#1}}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Applicative Regular Expressions using the Free Alternative},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
% Make links footnotes instead of hotlinks:
\renewcommand{\href}[2]{#2\footnote{\url{#1}}}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\title{Applicative Regular Expressions using the Free Alternative}

\begin{document}
\maketitle

\% Justin Le \% April 8, 2019

\emph{Originally posted on
\textbf{\href{https://blog.jle.im/entry/free-alternative-regexp.html}{in
Code}}.}

Today, we're going to implement applicative regular expressions and parsers (in
the style of the
\href{https://hackage.haskell.org/package/regex-applicative}{regex-applicative}
library) using free structures!

Free structures are some of my favorite tools in Haskell, and I've actually
written a few posts about them before, including
\href{https://blog.jle.im/entry/alchemical-groups.html}{this one using free
groups}, \href{https://blog.jle.im/entry/interpreters-a-la-carte-duet.html}{this
one on a free monad variation}, and
\href{https://blog.jle.im/entry/const-applicative-and-monoids.html}{this one on
a ``free'' applicative on a monoid}.

Regular expressions (and parsers) are ubiquitous in computer science and
programming, and I hope that demonstrating that they are pretty straightforward
to implement using free structures will help you see the value in free
structures without getting too bogged down in the details!

All of the code in this post is
\href{https://github.com/mstksg/inCode/tree/master/code-samples/misc/regexp.hs}{available
online} as a ``stack executable''. When you run it (\texttt{./regexp.hs}),
you'll load up a \emph{ghci} session with all of the definitions in scope, so
you can play around with the functions and types :)

This post should be accessible to late beginner or early intermediate Haskell
users, and requires some basic familiarity with pattern matching, algebraic data
types, and abstractions like \texttt{Monoid} and \texttt{Functor}, and do
notation.

\section{Regular Languages}\label{regular-languages}

A \emph{regular expression} is something that defines a \emph{regular language}.
\href{https://en.wikipedia.org/wiki/Regular_expression\#Formal_language_theory}{Formally},
it consists of the following base elements:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The empty set, which always fails to match.
\item
  The empty string, which always succeeds matching the empty string.
\item
  The literal character, denoting a single matching character
\end{enumerate}

And the following operations:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Concatenation: \texttt{RS}, sequence one after the other. A set product.
\item
  Alternation: \texttt{R\textbar{}S}, one or the other. A set union.
\item
  Kleene Star: \texttt{R*}, the repetition of \texttt{R} zero or more times.
\end{enumerate}

And that's \emph{all} that's in a regular expression. Nothing more, nothing
less. From these basic tools, you can derive the rest of the regexp operations
--- for example, \texttt{a+} can be expressed as \texttt{aa*}, and categories
like \texttt{\textbackslash{}w} can be expressed as alternations of valid
characters.

\subsection{Alternative}\label{alternative}

Looking at this, does this look a little familiar? It reminds me a lot of the
\texttt{Alternative} typeclass. If a functor \texttt{f} has an
\texttt{Alternative} instance, it means that it has:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{empty}, the failing operation
\item
  \texttt{pure\ x}, the always-succeeding operation (from the
  \texttt{Applicative} class)
\item
  \texttt{\textless{}*\textgreater{}}, the sequencing operation (from the
  \texttt{Applicative} class)
\item
  \texttt{\textless{}\textbar{}\textgreater{}}, the alternating operation
\item
  \texttt{many}, the ``zero or more'' operation.
\end{enumerate}

This\ldots looks a lot like the construction of a regular language, doesn't it?
It's almost as if \texttt{Alternative} has almost \emph{exactly} what we need.
The only thing missing is the literal character primitive.

If you're unfamiliar with \texttt{Alternative}, the
\href{https://wiki.haskell.org/Typeclassopedia}{typeclassopedia} has a good
step-by-step introduction. But for the purposes of this article, it's basically
just a ``double monoid'', with two ``combining'' actions
\texttt{\textless{}*\textgreater{}} and
\texttt{\textless{}\textbar{}\textgreater{}}, which roughly correspond to
\texttt{*} and \texttt{+} in the integers. It's basically pretty much nothing
more than 1-5 in the list above, and some distributivity laws.

So, one way we can look at regular expressions is ``The entire
\texttt{Alternative} interface, plus a character primitive''. \emph{But!}
There's another way of looking at this, that leads us directly to free
structures.

Instead of seeing things as ``\texttt{Alternative} with a character primitive'',
we can look at it as \emph{a character primitive enriched with a Alternative
instance}.

\section{Free}\label{free}

Let's write this out. Our character primitive will be:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L17{-}L18}

\KeywordTok{data} \DataTypeTok{Prim}\NormalTok{ a }\OtherTok{=} \DataTypeTok{Prim} \DataTypeTok{Char}\NormalTok{ a}
  \KeywordTok{deriving} \DataTypeTok{Functor}
\end{Highlighting}
\end{Shaded}

Note that because we're working with functors, applicatives, alternatives, etc.,
all of our regular expressions can have an associated ``result''. This is
because our regexp values will have a type parameter (which is required for
\texttt{Functor}, \texttt{Applicative}, and \texttt{Alternative}). We can choose
to ignore this type parameter, but we can also have some fun by using it to
represent a ``result'' that a regexp match will be interpreted as. This is
similar to the idea of ``capturing'' in industrial regexp applications.

Here, the value
\texttt{Prim\ \textquotesingle{}a\textquotesingle{}\ 1\ ::\ Prim\ Int} will
represent a primitive that matches on the character \texttt{a}, interpreting it
with a result of \texttt{1}.

And now\ldots we give it \texttt{Alternative} structure using the \emph{Free
Alternative}, from the
\emph{\href{https://hackage.haskell.org/package/free}{free}} package:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{import} \DataTypeTok{Control.Alternative.Free}

\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L20{-}L20}

\KeywordTok{type} \DataTypeTok{RegExp} \OtherTok{=} \DataTypeTok{Alt} \DataTypeTok{Prim}
\end{Highlighting}
\end{Shaded}

And that's it! That's our entire regular expression type! By giving a
\texttt{Alt} a \texttt{Functor}, we get all of the operations of
\texttt{Applicative} and \texttt{Alternative} over our base. That's because we
have \texttt{instance\ Applicative\ (Alt\ f)} and
\texttt{instance\ Alternative\ (Alt\ f)}. We now have:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The empty set, coming from \texttt{empty} from \texttt{Alternative}
\item
  The empty string, coming from \texttt{pure} from \texttt{Applicative}
\item
  The character primitive, coming from the underlying functor \texttt{Prim} that
  we are enhancing
\item
  The concatenation operation, from \texttt{\textless{}*\textgreater{}}, from
  \texttt{Applicative}.
\item
  The alternating operation, from \texttt{\textless{}\textbar{}\textgreater{}},
  from \texttt{Alternative}.
\item
  The kleene star, from \texttt{many}, from \texttt{Alternative}.
\end{enumerate}

All of these additions to our primitive come ``for free''!

Essentially, what a free structure gives us is the structure of the abstraction
(\texttt{Alternative}, here) automatically for our base type, and \emph{nothing
else}.

Remember that regular expressions have these operations, \emph{and nothing else}
--- no more, no less. That's exactly what the free Alternative gives us: these
operations and the primitive. No more, no less.

After adding some convenient wrappers\ldots we're done here!

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L22{-}L33}

\CommentTok{{-}{-} | charAs: Parse a given character as a given constant result.}
\OtherTok{charAs ::} \DataTypeTok{Char} \OtherTok{{-}\textgreater{}}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{RegExp}\NormalTok{ a}
\NormalTok{charAs c x }\OtherTok{=}\NormalTok{ liftAlt (}\DataTypeTok{Prim}\NormalTok{ c x)     }\CommentTok{{-}{-} liftAlt lets us use the underlying}
                                    \CommentTok{{-}{-} functor Prim in RegExp}

\CommentTok{{-}{-} | char: Parse a given character as itself.}
\OtherTok{char ::} \DataTypeTok{Char} \OtherTok{{-}\textgreater{}} \DataTypeTok{RegExp} \DataTypeTok{Char}
\NormalTok{char c }\OtherTok{=}\NormalTok{ charAs c c}

\CommentTok{{-}{-} | string: Parse a given string as itself.}
\OtherTok{string ::} \DataTypeTok{String} \OtherTok{{-}\textgreater{}} \DataTypeTok{RegExp} \DataTypeTok{String}
\NormalTok{string }\OtherTok{=} \FunctionTok{traverse}\NormalTok{ char              }\CommentTok{{-}{-} neat, huh}
\end{Highlighting}
\end{Shaded}

\subsection{Examples}\label{examples}

Let's try it out! Let's match on \texttt{(a\textbar{}b)(cd)*e} and return
\texttt{()}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L35{-}L38}

\OtherTok{testRegExp\_ ::} \DataTypeTok{RegExp}\NormalTok{ ()}
\NormalTok{testRegExp\_ }\OtherTok{=}\NormalTok{ void }\OperatorTok{$}\NormalTok{ (char }\CharTok{\textquotesingle{}a\textquotesingle{}} \OperatorTok{\textless{}|\textgreater{}}\NormalTok{ char }\CharTok{\textquotesingle{}b\textquotesingle{}}\NormalTok{)}
                  \OperatorTok{*\textgreater{}}\NormalTok{ many (string }\StringTok{"cd"}\NormalTok{)}
                  \OperatorTok{*\textgreater{}}\NormalTok{ char }\CharTok{\textquotesingle{}e\textquotesingle{}}
\end{Highlighting}
\end{Shaded}

\texttt{void} from \emph{Data.Functor} discards the results, since we only care
if it matches or not. But we use \texttt{\textless{}\textbar{}\textgreater{}}
and \texttt{*\textgreater{}} and \texttt{many} exactly how we'd expect to
concatenate and alternate things with \texttt{Applicative} and
\texttt{Alternative}.

Or maybe more interesting (but slightly more complicated), let's match on the
same pattern and return how many \texttt{cd}s are repeated

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L40{-}L43}

\OtherTok{testRegExp ::} \DataTypeTok{RegExp} \DataTypeTok{Int}
\NormalTok{testRegExp }\OtherTok{=}\NormalTok{ (char }\CharTok{\textquotesingle{}a\textquotesingle{}} \OperatorTok{\textless{}|\textgreater{}}\NormalTok{ char }\CharTok{\textquotesingle{}b\textquotesingle{}}\NormalTok{)}
          \OperatorTok{*\textgreater{}}\NormalTok{ (}\FunctionTok{length} \OperatorTok{\textless{}$\textgreater{}}\NormalTok{ many (string }\StringTok{"cd"}\NormalTok{))}
          \OperatorTok{\textless{}*}\NormalTok{ char }\CharTok{\textquotesingle{}e\textquotesingle{}}
\end{Highlighting}
\end{Shaded}

This one does require a little more finesse with \texttt{*\textgreater{}} and
\texttt{\textless{}*}: the arrows point towards which result to ``keep''. And
since \texttt{many\ (string\ "cd")\ ::\ RegExp\ {[}String{]}} (it returns a
list, with an item for each repetition), we can \texttt{fmap\ length} to get the
\texttt{Int} result of ``how many repetitions''.

However, we can also turn on \emph{-XApplicativeDo} and write it using do
notation, which requires a little less thought:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L45{-}L50}

\OtherTok{testRegExpDo ::} \DataTypeTok{RegExp} \DataTypeTok{Int}
\NormalTok{testRegExpDo }\OtherTok{=} \KeywordTok{do}
\NormalTok{    char }\CharTok{\textquotesingle{}a\textquotesingle{}} \OperatorTok{\textless{}|\textgreater{}}\NormalTok{ char }\CharTok{\textquotesingle{}b\textquotesingle{}}
\NormalTok{    cds }\OtherTok{\textless{}{-}}\NormalTok{ many (string }\StringTok{"cd"}\NormalTok{)}
\NormalTok{    char }\CharTok{\textquotesingle{}e\textquotesingle{}}
    \FunctionTok{pure}\NormalTok{ (}\FunctionTok{length}\NormalTok{ cds)}
\end{Highlighting}
\end{Shaded}

It's all a little bit like how we often use ``captures'' in regular expressions
to access a \emph{specific part} of a match. Here's an example in ruby:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{irb}\KeywordTok{\textgreater{}} \SpecialStringTok{/(a|b)((cd)*)e/}\AttributeTok{.match}\NormalTok{(}\StringTok{"acdcdcdcde"}\NormalTok{)}\KeywordTok{[}\DecValTok{2}\KeywordTok{]}
\KeywordTok{=\textgreater{}} \StringTok{"cdcdcdcd"}
\end{Highlighting}
\end{Shaded}

except we also include a ``post-processing'' process to get the length of the
number of repetitions.

Here's another handy regexp that matches on a digit between 0 to 9, and the
result is the digit \texttt{Int} itself:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L52{-}L53}

\OtherTok{digit ::} \DataTypeTok{RegExp} \DataTypeTok{Int}
\NormalTok{digit }\OtherTok{=}\NormalTok{ asum [ charAs (}\FunctionTok{intToDigit}\NormalTok{ i) i }\OperatorTok{|}\NormalTok{ i }\OtherTok{\textless{}{-}}\NormalTok{ [}\DecValTok{0}\OperatorTok{..}\DecValTok{9}\NormalTok{] ]}
\end{Highlighting}
\end{Shaded}

Here,
\texttt{asum\ {[}x,y,z{]}\ =\ x\ \textless{}\textbar{}\textgreater{}\ y\ \textless{}\textbar{}\textgreater{}\ z}:
it represents a choice between the items in a list.

We can again do some fancy things with it, like make a regexp
\texttt{\textbackslash{}{[}\textbackslash{}d\textbackslash{}{]}} that matches on
a digit inside \texttt{{[}} / \texttt{{]}} brackets:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L55{-}L56}

\OtherTok{bracketDigit ::} \DataTypeTok{RegExp} \DataTypeTok{Int}
\NormalTok{bracketDigit }\OtherTok{=}\NormalTok{ char }\CharTok{\textquotesingle{}[\textquotesingle{}} \OperatorTok{*\textgreater{}}\NormalTok{ digit }\OperatorTok{\textless{}*}\NormalTok{ char }\CharTok{\textquotesingle{}]\textquotesingle{}}
\end{Highlighting}
\end{Shaded}

\section{Parsing}\label{parsing}

Okay, so all we did was define a data structure that supports character
matching, concatenation, alternation, and starring. Big whoop. What we really
want to do is use it to parse things, right? How does the Free Alternative help
us with \emph{that}?

Well, it does a lot, actually. Let's look at two ways of doing this!

\subsection{Offloading Alternative}\label{offloading-alternative}

\subsubsection{What is Freeness?}\label{what-is-freeness}

The ``canonical'' way of using a free structure is by using its ``folding''
operation into a concrete structure with the proper instances. For example,
\texttt{foldMap} will turn a free monoid into a value of any monoid instance:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{foldMap}\OtherTok{ ::} \DataTypeTok{Monoid}\NormalTok{ m }\OtherTok{=\textgreater{}}\NormalTok{ (a }\OtherTok{{-}\textgreater{}}\NormalTok{ m) }\OtherTok{{-}\textgreater{}}\NormalTok{ ([a] }\OtherTok{{-}\textgreater{}}\NormalTok{ m)}
\end{Highlighting}
\end{Shaded}

\texttt{foldMap} lifts an \texttt{a\ -\textgreater{}\ m} into a
\texttt{{[}a{]}\ -\textgreater{}\ m} (or,
\texttt{FreeMonoid\ a\ -\textgreater{}\ m}), with a concrete monoid \texttt{m}.
The general idea is that using a free structure can ``defer'' the concretization
from between the time of construction to the time of use.

For example, we can construct value in the free monoid made from integers:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} | Lift the "primitive" \textasciigrave{}Int\textasciigrave{} into a value in the free monoid on \textasciigrave{}Int\textasciigrave{}.}
\CommentTok{{-}{-}}
\CommentTok{{-}{-} Analogous to \textasciigrave{}liftAlt\textasciigrave{} from earlier.}
\OtherTok{liftFM ::} \DataTypeTok{Int} \OtherTok{{-}\textgreater{}}\NormalTok{ [}\DataTypeTok{Int}\NormalTok{]}
\NormalTok{liftFM x }\OtherTok{=}\NormalTok{ [x]}

\OtherTok{myMon ::}\NormalTok{ [}\DataTypeTok{Int}\NormalTok{]}
\NormalTok{myMon }\OtherTok{=}\NormalTok{ liftFM }\DecValTok{1} \OperatorTok{\textless{}\textgreater{}}\NormalTok{ liftFM }\DecValTok{2} \OperatorTok{\textless{}\textgreater{}}\NormalTok{ liftFM }\DecValTok{3} \OperatorTok{\textless{}\textgreater{}}\NormalTok{ liftFM }\DecValTok{4}
\end{Highlighting}
\end{Shaded}

And now we can decide how we want to interpret
\texttt{\textless{}\textgreater{}} --- should it be \texttt{+}?

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\OperatorTok{\textgreater{}} \FunctionTok{foldMap} \DataTypeTok{Sum}\NormalTok{ myMon}
\DataTypeTok{Sum} \DecValTok{10}              \CommentTok{{-}{-} 1 + 2 + 3 + 4}
\end{Highlighting}
\end{Shaded}

Or should it be \texttt{*}?

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\OperatorTok{\textgreater{}} \FunctionTok{foldMap} \DataTypeTok{Product}\NormalTok{ myMon}
\DataTypeTok{Product} \DecValTok{24}          \CommentTok{{-}{-} 1 * 2 * 3 * 4}
\end{Highlighting}
\end{Shaded}

Or maybe even \texttt{max}?

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\OperatorTok{\textgreater{}} \FunctionTok{foldMap} \DataTypeTok{Max}\NormalTok{ myMon}
\DataTypeTok{Max} \DecValTok{4}          \CommentTok{{-}{-} 1 \textasciigrave{}max\textasciigrave{} 2 \textasciigrave{}max\textasciigrave{} 3 \textasciigrave{}max\textasciigrave{} 4}
\end{Highlighting}
\end{Shaded}

The idea is that we can ``defer'' the choice of concrete \texttt{Monoid} that
\texttt{\textless{}\textgreater{}} is interpreted under by first pushing 1, 2,
3, and 4 into a free monoid value. The free monoid on \texttt{Int} gives
\emph{exactly enough structure} to \texttt{Int} to do this job: no more, no
less.

To use \texttt{foldMap}, we say ``how to handle the base type'', and it lets us
handle the free structure in its entirety by offloading the behavior of
\texttt{\textless{}\textgreater{}} to the concrete monoid.

\subsubsection{Interpreting in State}\label{interpreting-in-state}

In practice, getting a result from a free structure is often about finding (or
creating) the right concrete \texttt{Alternative} that gives us the behavior we
want. In this case, we're in luck. There's a concrete \texttt{Alternative}
instance that works just the way we want: \texttt{StateT\ String\ Maybe}:

\begin{itemize}
\tightlist
\item
  Its \texttt{\textless{}*\textgreater{}} works by sequencing changes in state;
  in this case, we'll consider the state as ``characters yet to be parsed'', so
  sequential parsing fits perfectly with \texttt{\textless{}*\textgreater{}}.
  That's because combining regexps sequentially can be thought of as statefully
  chomping down on a string.
\item
  Its \texttt{\textless{}\textbar{}\textgreater{}} works by backtracking and
  trying again if it runs into a failure. It saves the state of the last
  successful point and resets to it on failure. This is exactly how we want
  regexp alternation \texttt{R\textbar{}S} to behave.
\end{itemize}

The ``folding'' operation of the free alternative is called \texttt{runAlt}:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{runAlt ::} \DataTypeTok{Alternative}\NormalTok{ f}
       \OtherTok{=\textgreater{}}\NormalTok{ (}\KeywordTok{forall}\NormalTok{ b}\OperatorTok{.}\NormalTok{ p b }\OtherTok{{-}\textgreater{}}\NormalTok{ f b)}
       \OtherTok{{-}\textgreater{}} \DataTypeTok{Alt}\NormalTok{ p a}
       \OtherTok{{-}\textgreater{}}\NormalTok{ f a}
\end{Highlighting}
\end{Shaded}

And in the case of \texttt{RegExp}, we have:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{runAlt ::} \DataTypeTok{Alternative}\NormalTok{ f}
       \OtherTok{=\textgreater{}}\NormalTok{ (}\KeywordTok{forall}\NormalTok{ b}\OperatorTok{.} \DataTypeTok{Prim}\NormalTok{ b }\OtherTok{{-}\textgreater{}}\NormalTok{ f b)}
       \OtherTok{{-}\textgreater{}} \DataTypeTok{RegExp}\NormalTok{ a}
       \OtherTok{{-}\textgreater{}}\NormalTok{ f a}
\end{Highlighting}
\end{Shaded}

If you're unfamiliar with the RankN type (the \texttt{forall\ b.} stuff),
there's a
\href{https://ocharles.org.uk/guest-posts/2014-12-18-rank-n-types.html}{nice
introduction here}. But basically, you just need to provide \texttt{runAlt} with
a function that can handle a \texttt{Prim\ b} for \emph{any} \texttt{b} (and not
just a specific one like \texttt{Int} or \texttt{Bool}).

So, like \texttt{foldMap}, we need to say ``how to handle our base type''. Here,
we have to answer ``How do we handle \texttt{Prim}?''

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L58{-}L63}

\OtherTok{processPrim ::} \DataTypeTok{Prim}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{StateT} \DataTypeTok{String} \DataTypeTok{Maybe}\NormalTok{ a}
\NormalTok{processPrim (}\DataTypeTok{Prim}\NormalTok{ c x) }\OtherTok{=} \KeywordTok{do}
\NormalTok{    d}\OperatorTok{:}\NormalTok{ds }\OtherTok{\textless{}{-}}\NormalTok{ get}
\NormalTok{    guard (c }\OperatorTok{==}\NormalTok{ d)}
\NormalTok{    put ds}
    \FunctionTok{pure}\NormalTok{ x}
\end{Highlighting}
\end{Shaded}

This lets us interpret a \texttt{Prim} as a \texttt{StateT\ String\ Maybe}
action where the state is the ``string left to be be processed''. Remember, a
\texttt{Prim\ a} contains the character we want to match on, and the \texttt{a}
value we want it to be interpreted as. To process a \texttt{Prim}, we:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Get the state's (the string left to be parsed) head and tail, using
  \texttt{get}. If this match fails, backtrack.
\item
  Using \texttt{guard}, backtrack unless the head matches what the \texttt{Prim}
  expects.
\item
  Set the state to be the original tail, using \texttt{put}. This is because we
  parsed the head already, so now the ``string left to be parsed'' is just the
  original tail.
\item
  The result is what the \texttt{Prim} says it should be.
\end{enumerate}

We can use this to write a function that matches the \texttt{RegExp} on a
prefix. We need to run the state action (using \texttt{evalStateT}) on the
string we want to parse:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L65{-}L66}

\OtherTok{matchPrefix ::} \DataTypeTok{RegExp}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{String} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a}
\NormalTok{matchPrefix re }\OtherTok{=}\NormalTok{ evalStateT (runAlt processPrim re)}
\end{Highlighting}
\end{Shaded}

And that's it! Our first solution:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchPrefix testRegexp\_ }\StringTok{"acdcdcde"}
\DataTypeTok{Just}\NormalTok{ ()}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchPrefix testRegexp\_ }\StringTok{"acdcdcdx"}
\DataTypeTok{Nothing}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchPrefix testRegexp }\StringTok{"acdcdcde"}
\DataTypeTok{Just} \DecValTok{3}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchPrefix testRegexp }\StringTok{"bcdcdcdcdcdcdcde"}
\DataTypeTok{Just} \DecValTok{7}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchPrefix digit }\StringTok{"9"}
\DataTypeTok{Just} \DecValTok{9}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchPrefix bracketDigit }\StringTok{"[2]"}
\DataTypeTok{Just} \DecValTok{2}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchPrefix (many bracketDigit) }\StringTok{"[2][3][4][5]"}
\DataTypeTok{Just}\NormalTok{ [}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{4}\NormalTok{,}\DecValTok{5}\NormalTok{]}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchPrefix (}\FunctionTok{sum} \OperatorTok{\textless{}$\textgreater{}}\NormalTok{ many bracketDigit) }\StringTok{"[2][3][4][5]"}
\DataTypeTok{Just} \DecValTok{14}
\end{Highlighting}
\end{Shaded}

\subsubsection{Wait, what just happened?}\label{wait-what-just-happened}

Okay, so that might have happened a little quicker than you expected. One minute
we were writing our primitive, and the next we had already finished. Here's the
entirety of the code, in a few lines of Haskell:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{type} \DataTypeTok{RegExp} \OtherTok{=} \DataTypeTok{Alt} \DataTypeTok{Prim}

\OtherTok{matchPrefix ::} \DataTypeTok{RegExp}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{String} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a}
\NormalTok{matchPrefix re }\OtherTok{=}\NormalTok{ evalStateT (runAlt processPrim re)}
  \KeywordTok{where}
\NormalTok{    processPrim (}\DataTypeTok{Prim}\NormalTok{ c x) }\OtherTok{=} \KeywordTok{do}
\NormalTok{      d}\OperatorTok{:}\NormalTok{ds }\OtherTok{\textless{}{-}}\NormalTok{ get}
\NormalTok{      guard (c }\OperatorTok{==}\NormalTok{ d)}
\NormalTok{      put ds}
      \FunctionTok{pure}\NormalTok{ x}
\end{Highlighting}
\end{Shaded}

And now we have a fully functioning regexp parser? What happened?

From a high-level view, remember that \texttt{Alt\ Prim} has, in its structure,
\texttt{pure}, \texttt{empty}, \texttt{Prim},
\texttt{\textless{}*\textgreater{}},
\texttt{\textless{}\textbar{}\textgreater{}}, and \texttt{many}\footnote{A
  caveat exists here for \texttt{many}. More on this later!}.

Essentially, what \texttt{runAlt} does is that it uses a given concrete
\texttt{Alternative} (here, \texttt{StateT\ String\ Maybe}) to get the behavior
of \texttt{pure}, \texttt{empty}, \texttt{\textless{}*\textgreater{}},
\texttt{\textless{}\textbar{}\textgreater{}}, and \texttt{many}. But! As we can
see from that list, \texttt{StateT} does \emph{not} have a built-in behavior for
\texttt{Prim}. And so, that's where \texttt{processPrim} comes in.

\begin{itemize}
\tightlist
\item
  For \texttt{Prim}, \texttt{runAlt} uses \texttt{processPrim}.
\item
  For \texttt{pure}, \texttt{empty}, \texttt{\textless{}*\textgreater{}},
  \texttt{\textless{}\textbar{}\textgreater{}}, and \texttt{many},
  \texttt{runAlt} uses \texttt{StateT\ String\ \ \ \ \ Maybe}'s
  \texttt{Alternative} instance.
\end{itemize}

So, really, 83\% of the work was done for us by \texttt{StateT}'s
\texttt{Alternative} instance, and the other 17\% is in \texttt{processPrim}.

Admittedly, this \emph{does} feel a little disappointing, or at least
anticlimactic. This makes us wonder: why even use \texttt{Alt} in the first
place? Why not just have \texttt{type\ RegExp\ =\ StateT\ String\ Maybe} and
write an appropriate
\texttt{char\ ::\ Char\ -\textgreater{}\ StateT\ String\ Maybe\ Char}? If
\texttt{StateT} does all of the work anyway, why even bother with \texttt{Alt},
the free Alternative?

One major advantage we get from using \texttt{Alt} is that \texttt{StateT}
is\ldots pretty powerful. It's actually \emph{stupid} powerful. It can represent
a lot of things\ldots most troubling, it can represent things that \emph{are not
regular expressions}. For example, something as simple as
\texttt{put\ "hello"\ ::\ StateT\ String\ Maybe\ ()} does not correspond to
\emph{any} regular expression.

So, while we can say that \texttt{Alt\ Prim} corresponds to ``regular
expressions, nothing less and nothing more'', we \emph{cannot} say the same
about \texttt{StateT\ String\ Maybe}.

\texttt{Alt\ Prim} contains a ``perfect fit'' representation of a regular
expression data type. Everything it can express is a regular expression, and
there is nothing it can express that \emph{isn't} a regular
expression.\footnote{Note that there are some caveats that should be noted here,
  due to laziness in Haskell. We will go deeper into this later.}

Here, we can think of \texttt{StateT} as the context that we use to
\emph{interpret} a \texttt{RegExp} as a \emph{parser}. But, there might be
\emph{other} ways we want to work with a \texttt{RegExp}. For example, we might
want to inspect it and ``print'' it out for inspection. This is something we
can't do with \texttt{StateT}.

We can't say that \texttt{StateT\ String\ Maybe} ``is'' a regular expression ---
only that it can represent a parser based on a regular expression. But we
\emph{can} say that about \texttt{Alt\ Prim}.

\subsection{Using the Free structure
directly}\label{using-the-free-structure-directly}

Alright, that's great and all. But what if we didn't want to offload 83\% of the
behavior to a type that has already been written for us. Is there a way we can
directly use the structure of \texttt{Alt} itself to write our parser?

This is analogous to asking, what if we wanted to actually write a function on a
list (by pattern matching on \texttt{:} and \texttt{{[}{]}}) instead of always
using \texttt{foldMap}? Can we directly operate on the structure of the list
instead of using \texttt{foldMap} with some concrete monoid?

I'm glad you asked! Let's look at the definition of the free alternative:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{newtype} \DataTypeTok{Alt}\NormalTok{ f a }\OtherTok{=} \DataTypeTok{Alt}\NormalTok{ \{}\OtherTok{ alternatives ::}\NormalTok{ [}\DataTypeTok{AltF}\NormalTok{ f a] \}}

\KeywordTok{data} \DataTypeTok{AltF}\NormalTok{ f a }\OtherTok{=} \KeywordTok{forall}\NormalTok{ r}\OperatorTok{.} \DataTypeTok{Ap}\NormalTok{ (f r) (}\DataTypeTok{Alt}\NormalTok{ f (r }\OtherTok{{-}\textgreater{}}\NormalTok{ a))}
              \OperatorTok{|}           \DataTypeTok{Pure}\NormalTok{ a}
\end{Highlighting}
\end{Shaded}

It's a mutually recursive (and
\href{https://twanvl.nl/blog/haskell/non-regular1}{non-regular}) type, so it
might be a little confusing. One way to understand \texttt{Alt} is that
\texttt{Alt\ xs} contains a \emph{list of alternatives}, or a list of
\texttt{\textless{}\textbar{}\textgreater{}}s. And each of those alternatives is
an \texttt{AltF}, which is a \emph{sequence of \texttt{f\ a}s} chained by
\texttt{\textless{}*\textgreater{}} (as a chain of function applications).

You can essentially think of \texttt{AltF\ f\ a} as a linked list
\texttt{{[}f\ r{]}}, except with a different \texttt{r} for each item.
\texttt{Ap} is cons (\texttt{:}), containing the \texttt{f\ r}, and
\texttt{Pure} is nil (\texttt{{[}{]}}). The \texttt{forall\ r.} here is
\emph{-XExistentialQuantification}, and is what lets us have a different
intermediate type for each item in our chain.

All in all, \texttt{Alt\ f} is like a list (\texttt{Alt} list) of lists
(\texttt{AltF} chains), which take turn alternating between alternative lists
and application sequences. A list of chains of lists of chains of lists of
chains \ldots{}

In the big picture, you can think of \texttt{Alt\ f} as a ``normalized'' form of
successive or nested \texttt{\textless{}*\textgreater{}} and
\texttt{\textless{}\textbar{}\textgreater{}}s, similar to how \texttt{{[}a{]}}
is a ``normalized'' form of successive \texttt{\textless{}\textgreater{}}s.

Ultimately we want to write a
\texttt{RegExp\ a\ -\textgreater{}\ String\ -\textgreater{}\ Maybe\ a}, which
parses a string based on a \texttt{RegExp}. We can do this using the most basic
of all Haskell function-writing techniques: pattern matching on each case, and
handling each case.

First, the top-level \texttt{Alt} case. When faced with a list of chains, we can
try to parse each one. The result is the first success.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L68{-}L69}

\OtherTok{matchAlts ::} \DataTypeTok{RegExp}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{String} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a}
\NormalTok{matchAlts (}\DataTypeTok{Alt}\NormalTok{ res) xs }\OtherTok{=}\NormalTok{ asum [ matchChain re xs }\OperatorTok{|}\NormalTok{ re }\OtherTok{\textless{}{-}}\NormalTok{ res ]}
\end{Highlighting}
\end{Shaded}

Here, \texttt{asum\ ::\ {[}Maybe\ a{]}\ -\textgreater{}\ Maybe\ a} finds the
first \texttt{Just} (success) in a list of attempts.

Now, we need to handle the chain case. To do this, we can pattern match on each
constructor, and handle each case.

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{matchChain ::} \DataTypeTok{AltF} \DataTypeTok{Prim}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{String} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a}
\NormalTok{matchChain (}\DataTypeTok{Ap}\NormalTok{ (}\DataTypeTok{Prim}\NormalTok{ c x) next) cs }\OtherTok{=}\NormalTok{ \_}
\NormalTok{matchChain (}\DataTypeTok{Pure}\NormalTok{ x)             cs }\OtherTok{=}\NormalTok{ \_}
\end{Highlighting}
\end{Shaded}

From here, it's mostly ``type tetris''! We just continually ask GHC what goes in
what holes (and what types need to change) until we get something that
typechecks.

In the end of the very mechanical process, we get:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L71{-}L76}

\OtherTok{matchChain ::} \DataTypeTok{AltF} \DataTypeTok{Prim}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{String} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a}
\NormalTok{matchChain (}\DataTypeTok{Ap}\NormalTok{ (}\DataTypeTok{Prim}\NormalTok{ c x) next) cs }\OtherTok{=} \KeywordTok{case}\NormalTok{ cs }\KeywordTok{of}
\NormalTok{    []  }\OtherTok{{-}\textgreater{}} \DataTypeTok{Nothing}
\NormalTok{    d}\OperatorTok{:}\NormalTok{ds }\OperatorTok{|}\NormalTok{ c }\OperatorTok{==}\NormalTok{ d    }\OtherTok{{-}\textgreater{}}\NormalTok{ matchAlts ((}\OperatorTok{$}\NormalTok{ x) }\OperatorTok{\textless{}$\textgreater{}}\NormalTok{ next) ds}
         \OperatorTok{|} \FunctionTok{otherwise} \OtherTok{{-}\textgreater{}} \DataTypeTok{Nothing}
\NormalTok{matchChain (}\DataTypeTok{Pure}\NormalTok{ x)             \_      }\OtherTok{=} \DataTypeTok{Just}\NormalTok{ x}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  If it's \texttt{Ap} (analogous to cons, \texttt{:}), it means we're in the
  middle of the chain.

  \begin{itemize}
  \tightlist
  \item
    If the input string is empty, then we fail to match.
  \item
    Otherwise, here's the interesting thing. We have the \texttt{Prim\ r} with
    the character we want to match, the first letter in the string, and
    \texttt{next\ \ \ \ \ ::\ RegExp\ (r\ -\textgreater{}\ a)}, the next
    \texttt{RegExp} in our sequential parsing chain.

    \begin{itemize}
    \tightlist
    \item
      If the match is a success, we continue down the chain, to \texttt{next}.
      We just need to massage the types a bit to make it all work out.
    \item
      Otherwise, it's a failure. We're done here.
    \end{itemize}
  \end{itemize}
\item
  If it's \texttt{Pure\ x} (analogous to nil, \texttt{{[}{]}}), it means we're
  at the end of the chain. We return the result in \texttt{Just}.
\end{enumerate}

In the end though, you don't really need to understand any of this \emph{too}
deeply in order to write this. Sure, it's nice to understand what \texttt{Ap},
\texttt{Pure}, \texttt{AltF}, etc. really ``mean''. But, we don't have to ---
the types take care of all of it for you :)

That should be good enough to implement another prefix parser:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchAlts testRegexp\_ }\StringTok{"acdcdcde"}
\DataTypeTok{Just}\NormalTok{ ()}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchAlts testRegexp\_ }\StringTok{"acdcdcdx"}
\DataTypeTok{Nothing}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchAlts testRegexp }\StringTok{"acdcdcde"}
\DataTypeTok{Just} \DecValTok{3}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchAlts testRegexp }\StringTok{"bcdcdcdcdcdcdcde"}
\DataTypeTok{Just} \DecValTok{7}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchAlts digit }\StringTok{"9"}
\DataTypeTok{Just} \DecValTok{9}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchAlts bracketDigit }\StringTok{"[2]"}
\DataTypeTok{Just} \DecValTok{2}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchAlts (many bracketDigit) }\StringTok{"[2][3][4][5]"}
\DataTypeTok{Just}\NormalTok{ [}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{4}\NormalTok{,}\DecValTok{5}\NormalTok{]}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ matchAlts (}\FunctionTok{sum} \OperatorTok{\textless{}$\textgreater{}}\NormalTok{ many bracketDigit) }\StringTok{"[2][3][4][5]"}
\DataTypeTok{Just} \DecValTok{14}
\end{Highlighting}
\end{Shaded}

\subsection{What did we do?}\label{what-did-we-do}

The two attempts here can be compared to using lists via \texttt{foldMap}
vs.~using lists via pattern matching.

Because lists act as a free monoid, \emph{any list function} can be written
using \texttt{foldMap} and not directly pattern matching. If this seems
unbelievable to you, try finding a function that can't --- you might be
surprised!

However, because lists are an algebraic data type, \emph{any list function} can
be written using straight up pattern matching on \texttt{:} and \texttt{{[}{]}}.

One nice thing about list is that, no matter how you assemble it, it always ends
up as a series of conses and nil. We say that the free monoid is
\emph{normalizing}. That is,
\texttt{{[}1,2,3{]}\ \textless{}\textgreater{}\ {[}4{]}} has the same
representation as
\texttt{{[}1{]}\ \textless{}\textgreater{}\ {[}2,3{]}\ \textless{}\textgreater{}\ {[}4{]}.}
When we pattern match on \texttt{:} and \texttt{{[}{]}}, we can't distinguish
between those two original methods of creation.

\texttt{Alt} is normalizing as well. An example of a possible variant that is
\emph{not} normalizing is:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{RegExp}\NormalTok{ a }\OtherTok{=} \DataTypeTok{Empty}
              \OperatorTok{|} \DataTypeTok{Pure}\NormalTok{ a}
              \OperatorTok{|} \DataTypeTok{Prim} \DataTypeTok{Char}\NormalTok{ a}
              \OperatorTok{|} \KeywordTok{forall}\NormalTok{ r}\OperatorTok{.} \DataTypeTok{Seq}\NormalTok{ (}\DataTypeTok{RegExp}\NormalTok{ r) (}\DataTypeTok{RegExp}\NormalTok{ (r }\OtherTok{{-}\textgreater{}}\NormalTok{ a))}
              \OperatorTok{|} \DataTypeTok{Union}\NormalTok{ (}\DataTypeTok{RegExp}\NormalTok{ a) (}\DataTypeTok{RegExp}\NormalTok{ a)}
              \OperatorTok{|} \DataTypeTok{Many}\NormalTok{ (}\DataTypeTok{RegExp}\NormalTok{ a)}
\end{Highlighting}
\end{Shaded}

This is how we \emph{might} have written \texttt{RegExp}, if we didn't know
about the free alternative. However, this representation is not normalizing,
because we have two \texttt{RegExp\ a} values that represent the same regexp:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} | a|(b|c)}
\OtherTok{abc1 ::} \DataTypeTok{RegExp} \DataTypeTok{Int}
\NormalTok{abc1 }\OtherTok{=} \DataTypeTok{Prim} \CharTok{\textquotesingle{}a\textquotesingle{}} \DecValTok{1} \OtherTok{\textasciigrave{}Union\textasciigrave{}}\NormalTok{ (}\DataTypeTok{Prim} \CharTok{\textquotesingle{}b\textquotesingle{}} \DecValTok{2} \OtherTok{\textasciigrave{}Union\textasciigrave{}} \DataTypeTok{Prim} \CharTok{\textquotesingle{}c\textquotesingle{}} \DecValTok{3}\NormalTok{)}

\CommentTok{{-}{-} | (a|b)|c}
\OtherTok{abc2 ::} \DataTypeTok{RegExp} \DataTypeTok{Int}
\NormalTok{abc2 }\OtherTok{=}\NormalTok{ (}\DataTypeTok{Prim} \CharTok{\textquotesingle{}a\textquotesingle{}} \DecValTok{1} \OtherTok{\textasciigrave{}Union\textasciigrave{}} \DataTypeTok{Prim} \CharTok{\textquotesingle{}b\textquotesingle{}} \DecValTok{2}\NormalTok{) }\OtherTok{\textasciigrave{}Union\textasciigrave{}} \DataTypeTok{Prim} \CharTok{\textquotesingle{}c\textquotesingle{}} \DecValTok{3}
\end{Highlighting}
\end{Shaded}

These two match the same thing. But they have different representations. This
representation \emph{not} normalizing, since the same regexp can be expressed in
two different ways.

\texttt{Alt\ Prim} is better because if two regexps are the same\ldots they will
correspond to the same \texttt{Alt\ Prim}. It forces each value to exist in a
``canonical'' normalized form.

This means that when we eventually write our parsing function
\texttt{matchAlts}, \emph{we aren't allowed to care} about ``how'' the regexps
are made. We aren't \emph{allowed} to distinguish between
\texttt{(a\textbar{}b)\textbar{}c} and \texttt{a\textbar{}(b\textbar{}c)}. The
normalizing property of \texttt{Alt} means that we are forced to treat both of
those the \emph{exact} same way. It \emph{forces} us to obey the laws about the
structures of regular languages.

It's easy to imagine a bug that might occur if we accidentally treated
\texttt{(a\textbar{}b)\textbar{}c} differently than
\texttt{a\textbar{}(b\textbar{}c)} --- and indeed it sounds like an easy bug to
accidentally make if we are using the non-normalizing representation.

Using \texttt{Alt} instead of rolling our own regexp expression not only
enforces our integrity, but it also eliminates a huge class of potential bugs.

However, it should be noted that, while \texttt{Alt\ f} is strongly normalizing
with respect to Alternative structure, \texttt{Alt\ Prim} isn't strongly
normalizing with respect to \emph{regular expressions} structure in every single
case. For example, \texttt{Alt\ Prim} will still treat \texttt{a\textbar{}a} as
different from \texttt{a}. This is mostly because \texttt{Alt} has to be
``agnostic'' to \texttt{f}. But, like with all structural ``type safety'', I
always follow this rule of thumb: \emph{A lot of safety is better than no
safety}. This method can't eliminate \emph{all} bugs arising from this angle,
but it can eliminate a whoooole lot.

\section{Some subtle caveats}\label{some-subtle-caveats}

Before we conclude, let's take some time to clarify a subtle point. Feel free to
skip this whole section if you don't care about the fact that these aren't
identical to the mathematical formalism of regular languages.

While we can \emph{use} \texttt{RegExp} just like a regular expression, the
formal concept of regular expressions is actually slightly different, due to one
pesky thing: \textbf{laziness}.

We really \emph{shouldn't} be too surprised, since laziness actually throws a
wrench into a \emph{lot} of Haskell abstractions that are based on math. For
example, laziness is the reason that lists aren't ``true'' mathematical free
monoids.

The reason is that because of laziness and unbounded recursion, we can create an
``infinite'' regular language:
\texttt{a\textbar{}aa\textbar{}aaa\textbar{}aaaa\textbar{}aaaaa\textbar{}aaaaaa\textbar{}...},
forever and ever. However, infinite regular expressions aren't allowed in the
mathematical version. In Haskell, unfortunately, there is no way to ``turn off''
recursion: we're stuck with it.

Even more unfortunately, this is actually how the \texttt{Alt} encoding of the
free alternative above implements \texttt{many}. \texttt{a*} is implemented as
\texttt{\textbar{}a\textbar{}aa\textbar{}aaa\textbar{}aaaa\textbar{}aaaaa\textbar{}...},
infinitely. So the representation actually \emph{relies} on laziness and
infinite recursion to do its job. If you look at the contents of
\texttt{many\ (char\ \textquotesingle{}a\textquotesingle{})}, you will see an
infinite list.

``Haskell without recursion'' is fine with \texttt{Alt} for a
``\href{https://en.wikipedia.org/wiki/Star-free_language}{star-free language}'',
but it won't cut it for a regular one.

For the purposes we talked about in this post, this doesn't matter. However,
this does create serious issues if we want to write a
\href{https://en.wikipedia.org/wiki/Nondeterministic_finite_automaton}{non-deterministic
finite automata based parser} (NFA), which is the de facto standard for
implementing fast regexp parsers. We can only really generate an NFA if we have
a \emph{finite} value, which means anything using \texttt{many} is out of the
question.

Not all hope is lost, however. We can actually use the ``final encoding'' of
\texttt{Alt}, from \emph{Control.Alternative.Free.Final}, to gain a
\texttt{many} that is non-recursive.

Using the final encoding means we lose the ``pattern match'' method, and can
only use the \texttt{runAlt} method. However, we can off-load to
\texttt{Alternative} instances that have non-recursive \texttt{many} (like the
\texttt{RE} type from \emph{regex-applicative}) that allows us to generate an
NFA parser. While this still has issues because Haskell allows general
recursion, at least \texttt{many} in specific is no longer dependent on infinite
structures.

There's another interesting point to be made, however, regarding compatibility
with NFAs. Even though this recursive encoding doesn't allow us to create an
\emph{explicit} NFA (a full graph with nodes and transitions), it does allow us
to make an \emph{implicit} one. We can't ever make an \emph{explicit} NFA
involving \texttt{many} because the naive \texttt{many} in the normal
\texttt{Alt} gives us an infinite data structure, so we get an infinite graph.

However, our implementation of \texttt{matchPrefix} is actually an
\emph{implicit} NFA in the GHC runtime, where the `states' can be considered
function pointers on the heap. These pointers refer to other pointers, and the
overall behavior works like an unoptimized NFA under the hood that's assembled
as we go. This circumvents the infinite data structure problem because GHC
Haskell internally implements recursion by cycles in pointer structures.

\section{Wrapping Up}\label{wrapping-up}

Let's wrap things up! As a cherry on top, we can write our final function to
find matches anywhere inside a string by using \texttt{tails} (which gives us
all prefixes in a string) and \texttt{mapMaybe} (which maps \texttt{matchPrefix}
on every prefix and keeps the successes). It's also useful to write a function
to get the \emph{first} successful match, using \texttt{listToMaybe}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} source: https://github.com/mstksg/inCode/tree/master/code{-}samples/misc/regexp.hs\#L78{-}L82}

\OtherTok{matches ::} \DataTypeTok{RegExp}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{String} \OtherTok{{-}\textgreater{}}\NormalTok{ [a]}
\NormalTok{matches re }\OtherTok{=}\NormalTok{ mapMaybe (matchPrefix re) }\OperatorTok{.}\NormalTok{ tails}

\OtherTok{firstMatch ::} \DataTypeTok{RegExp}\NormalTok{ a }\OtherTok{{-}\textgreater{}} \DataTypeTok{String} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a}
\NormalTok{firstMatch re }\OtherTok{=}\NormalTok{ listToMaybe }\OperatorTok{.}\NormalTok{ matches re}
\end{Highlighting}
\end{Shaded}

This is pretty efficient, due to how \texttt{matchPrefix} short-circuits with
\texttt{Nothing} as soon as it fails, and how \texttt{listToMaybe}
short-circuits as soon as it finds a \texttt{Just}.

Hopefully from this, you can see the value of free structures :)

\begin{itemize}
\tightlist
\item
  Given some base primitive, they give you exactly the structure you need --- no
  more, no less.
\item
  They let you work with your type safely, and then unsafely ``run'' it inside
  different useful contexts.
\item
  They are normalizing, so you are not allowed to make ``illegal'' distinctions.
  This eliminates a class of bugs where you accidentally treat two cases
  differently when they are meant to be treated the same way.
\end{itemize}

Our journey from going to regular languages to \texttt{Alt\ Prim} was to
recognize that the structures involved in regular expressions matched an
\texttt{Alternative} interface ``plus'' some extra primitives --- and then
shifting our perspective to enriching a primitive with \texttt{Alternative}
energy.

Where can we go from here? First, try playing around with the
\href{https://github.com/mstksg/inCode/tree/master/code-samples/misc/regexp.hs}{sample
code}. One easy addition would be to add other types of primitives:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Prim}\NormalTok{ a }\OtherTok{=}
    \DataTypeTok{Only} \DataTypeTok{Char}\NormalTok{ a                 }\CommentTok{{-}{-} \^{} match a char with a given result}
  \OperatorTok{|} \DataTypeTok{Letter}\NormalTok{ a                    }\CommentTok{{-}{-} \^{} match any letter with the same result}
  \OperatorTok{|} \DataTypeTok{Digit}\NormalTok{    (}\DataTypeTok{Int}  \OtherTok{{-}\textgreater{}}\NormalTok{ a)        }\CommentTok{{-}{-} \^{} match any digit, with a result computed from it}
  \OperatorTok{|} \DataTypeTok{Wildcard}\NormalTok{ (}\DataTypeTok{Char} \OtherTok{{-}\textgreater{}}\NormalTok{ a)        }\CommentTok{{-}{-} \^{} match any char, with a computed result}
  \OperatorTok{|} \DataTypeTok{Satisfy}\NormalTok{ (}\DataTypeTok{Char} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a)   }\CommentTok{{-}{-} \^{} match potentially any char, based on a function}
\end{Highlighting}
\end{Shaded}

so we can support a lot of the basic character classes that many implementations
of regular expressions support. Try this out in the
\href{https://github.com/mstksg/inCode/tree/master/code-samples/misc/regexp.hs}{sample
code} as an exercise!

One fun thing you can do also is to use our regexp type to generate a string
that it would match on. Try doing this both in the \texttt{runAlt}-based method
and also the explicit pattern matching method!

Another interesting direction we can take, along the lines of
\href{https://www.microsoft.com/en-us/research/publication/build-systems-la-carte/}{build
systems a la carte}, is experimenting with different free structures to give
rise to different types of languages/expressions. For example, if we use the
free \emph{Applicative}, we get a language that has only concatenation and empty
strings and primitives, and no alternations. It's like regular expressions with
no \texttt{\textbar{}}, or basically only straight up matches. If we use the
free \emph{Monad}, we get a context-sensitive language with no backtracking. If
we use the free \emph{MonadPlus}, we get a context-sensitive language with
backtracking. And if we use the (redundant) free \emph{Functor}\ldots we get a
language that can parse a string of one and only one character. It's nice that
we get this sort of ``a la carte'' scaling system by our choice of free
structure.

I hope that after working through this example, you will begin to start
recognizing opportunities for using free structures everywhere you look! Once
you start, it's hard to stop :)

\section{Special Thanks}\label{special-thanks}

I am very humbled to be supported by an amazing community, who make it possible
for me to devote time to researching and writing these posts. Very special
thanks to my supporters at the ``Amazing'' level on
\href{https://www.patreon.com/justinle/overview}{patreon}, Sam Stites and Josh
Vera! :)

\section{Signoff}\label{signoff}

Hi, thanks for reading! You can reach me via email at
\href{mailto:justin@jle.im}{\nolinkurl{justin@jle.im}}, or at twitter at
\href{https://twitter.com/mstk}{@mstk}! This post and all others are published
under the \href{https://creativecommons.org/licenses/by-nc-nd/3.0/}{CC-BY-NC-ND
3.0} license. Corrections and edits via pull request are welcome and encouraged
at \href{https://github.com/mstksg/inCode}{the source repository}.

If you feel inclined, or this post was particularly helpful for you, why not
consider \href{https://www.patreon.com/justinle/overview}{supporting me on
Patreon}, or a \href{bitcoin:3D7rmAYgbDnp4gp4rf22THsGt74fNucPDU}{BTC donation}?
:)

\end{document}
