\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{\textbf{#1}}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Functors to Monads: A Story of Shapes},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
% Make links footnotes instead of hotlinks:
\renewcommand{\href}[2]{#2\footnote{\url{#1}}}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\title{Functors to Monads: A Story of Shapes}

\begin{document}
\maketitle

\% Justin Le \% November 4, 2024

\emph{Originally posted on
\textbf{\href{https://blog.jle.im/entry/functors-to-monads-a-story-of-shapes.html}{in
Code}}.}

For many years now I've been using a mental model and intuition that has guided
me well for understanding and teaching and using functors, applicatives, monads,
and other related Haskell abstractions, as well as for approaching learning new
ones. Sometimes when teaching Haskell I talk about this concept and assume
everyone already has heard it, but I realize that it's something universal yet
easy to miss depending on how you're learning it. So, here it is: how I
understand the Functor and other related abstractions and free constructions in
Haskell.

The crux is this: instead of thinking about what \texttt{fmap} changes, ask:
what does \texttt{fmap} keep \emph{constant}?

This isn't a rigorous understanding and isn't going to explain \emph{every}
aspect about \emph{every} Functor, and will probably only be useful if you
already know a little bit about Functors in Haskell. But it's a nice intuition
trick that has yet to majorly mislead me.

\section{The Secret of Functors}\label{the-secret-of-functors}

First of all, \emph{what is a Functor}? A capital-F Functor, that is, the
Haskell typeclass and abstraction. Ask a random Haskeller on the street and
they'll tell you that it's something that can be ``mapped over'', like a list or
an optional. Maybe some of those random Haskellers will feel compelled to
mention that this mapping should follow some laws\ldots they might even list the
laws. Ask them why these laws are so important and maybe you'll spend a bit of
time on this rhetorical street of Haskellers before finding one confident enough
to give an answer.

So I'm going to make a bit of a tautological leap: a \emph{Functor} gives you a
way to ``map over'' values in a way that \emph{preserves shape}. And what is
``shape''? A shape is \emph{the thing that fmap preserves}.

The Functor typeclass is simple enough: for \texttt{Functor\ f}, you have a
function
\texttt{fmap\ ::\ (a\ -\textgreater{}\ b)\ -\textgreater{}\ f\ a\ -\textgreater{}\ f\ b},
along with \texttt{fmap\ id\ =\ id} and
\texttt{fmap\ f\ .\ fmap\ g\ =\ fmap\ (f\ .\ g)}. Cute things you can drop into
quickcheck to prove for your instance, but it seems like those laws are hiding
some sort of deeper, fundamental truth.

The more Functors you learn about, the more you see that \texttt{fmap} seems to
always preserve ``something'':

\begin{itemize}
\tightlist
\item
  For lists, \texttt{fmap} preserves length and relative orderings.
\item
  For optionals (\texttt{Maybe}), \texttt{fmap} preserves \emph{presence} (the
  fact that something is there or not). It cannot flip a \texttt{Just} to a
  \texttt{Nothing} or vice versa.
\item
  For \texttt{Either\ e}, \texttt{fmap} preserves \emph{the error} (if it
  exists) or the fact that it was succesful.
\item
  For \texttt{Map\ k}, \texttt{fmap} preserves \emph{the keys}: which keys
  exist, how many there are, their relative orderings, etc.
\item
  For \texttt{IO}, \texttt{fmap} preserves \emph{the IO effect}. Every bit of
  external I/O that an IO action represents is unchanged by an \texttt{fmap}, as
  well as exceptions.
\item
  For \texttt{Writer\ w} or \texttt{(,)\ w}, \texttt{fmap} preserves the
  ``logged'' \texttt{w} value, leaving it unchanged. Same for \texttt{Const\ w}.
\item
  For \texttt{Tree}, \texttt{fmap} preserves \emph{the tree structure}: how many
  layers, how big they are, how deep they are, etc.
\item
  For \texttt{State\ s}, \texttt{fmap} preserves what happens to the input state
  \texttt{s}. How a \texttt{State\ s} transform a state value \texttt{s} is
  unchanged by \texttt{fmap}
\item
  For \texttt{ConduitT\ i\ o\ m} from
  \emph{\href{https://hackage.haskell.org/package/conduit}{conduit}},
  \texttt{fmap} preserves what the conduit pulls upstream and what it yields
  downstream. \texttt{fmap} will not cause the conduit to yield more or
  different objects, nor cause it to consume/pull more or less.
\item
  For parser-combinator \texttt{Parser}, \texttt{fmap} preserves what input is
  consumed or would fail to be consumed. \texttt{fmap} cannot change whether an
  input string would fail or succeed, and it cannot change how much it consumes.
\item
  For
  \emph{\href{https://hackage.haskell.org/package/optparse-applicative}{optparse-applicative}}
  \texttt{Parser}s, \texttt{fmap} preserves the command line arguments
  available. It leaves the \texttt{-\/-help} message of your program unchanged.
\end{itemize}

It seems like as soon as you define a \texttt{Functor} instance, or as soon as
you find out that some type has a \texttt{Functor} instance, it magically
induces some sort of \ldots{} ``thing'' that must be preserved.\footnote{There
  are \emph{some} exceptions, especially degenerate cases like
  \texttt{Writer\ ()} aka \texttt{Identity} which add no meaningful structure.
  So for these this mental model isn't that useful.} A \emph{conserved quantity
must exist}. It reminds me a bit of
\href{https://en.wikipedia.org/wiki/Noether\%27s_theorem}{Noether's Theorem} in
Physics, where any continuous symmetry ``induces'' a conserved quantity (like
how translation symmetry ``causes'' conservation of momentum). In Haskell, every
lawful \texttt{Functor} instance induces a conserved quantity. I don't know if
there is a canonical name for this conserved quantity, but I like to call it
``shape''.

\section{A Story of Shapes}\label{a-story-of-shapes}

The word ``shape'' is chosen to be as devoid of external baggage/meaning as
possible while still having \emph{some}. The word isn't important as much as
saying that there is \emph{some} ``thing'' preserved by \texttt{fmap}, and not
exactly the nature of that ``thing''. The \emph{nature} of that thing changes a
lot from Functor to Functor, where we might better call it an ``effect'' or a
``structure'' specifically, but that \emph{some} ``thing'' exists is almost
universal.

Of course, the value if this ``thing'' having a canonical name at all is
debatable. I were to coin a completely new term I might call it a ``conserved
charge'' or ``gauge'' in allusion to physics. But the most useful name probably
would be shape.

For some \texttt{Functor} instances, the word shape is more literal than others.
For trees, for instance, you have the literal shape of the tree preserved. For
lists, the ``length'' could be considered a literal shape. \texttt{Map\ k}'s
shape is also fairly literal: it describes the structure of keys that exist in
the map. But for \texttt{Writer\ w} and \texttt{Const\ w}, shape can be
interpreted as some information outside of the values you are mapping that is
left unchanged by mapping. For \texttt{Maybe} and \texttt{Either\ e} shape also
considers if there has been any short-circuiting. For \texttt{State\ s} and
\texttt{IO} and \texttt{Parser}, ``shape'' involves some sort of
side-computation or consumption that is left unchanged by \texttt{fmap}, often
called an effect. For \emph{optparse-applicative}, ``shape'' involves some sort
of inspectable and observable static aspects of a program. ``Shape'' comes in
all forms.

But, this intuition of ``looking for that conserved quantity'' is very helpful
for learning \emph{new} Functors. If you stumble onto a new type that you know
is a \texttt{Functor} instance, you can immediately ask ``What \emph{shape} is
this \texttt{fmap} preserving?'', and it will almost always yield insight into
that type.

This viewpoint also sheds insight onto why \texttt{Set.map} isn't a good
candidate for \texttt{fmap} for
\emph{\href{https://hackage.haskell.org/package/containers/docs/Data-Set.html}{Data.Set}}:
What ``thing'' does \texttt{Set.map\ f} preserve? Not size, for sure. In a
hypothetical world where we had
\texttt{ordfmap\ ::\ Ord\ b\ =\textgreater{}\ (a\ -\textgreater{}\ b)\ -\textgreater{}\ f\ a\ -\textgreater{}\ f\ b},
we would still need \texttt{Set.map} to preserve \emph{something} for it to be
useful as an ``Ord-restricted Functor''.\footnote{Incidentally, \texttt{Set.map}
  \emph{does} preserve one thing: non-emptiness. You can't \texttt{Set.map} an
  empty set into a non-empty one and vice versa. So, maybe if we
  recontextualized \texttt{Set} as a ``search for at least one result''
  \texttt{Functor} or \texttt{Monad} where you could only ever observe a single
  value, \texttt{Set.map} would work for Ord-restricted versions of those
  abstractions, assuming lawful \texttt{Ord} instances.}

\subsection{A Result}\label{a-result}

Before we move on, let's look at another related and vague concept that is
commonly used when discussing functors: \texttt{fmap} is a way to map a function
that \emph{preserves the shape} and \emph{changes the result}.

If \emph{shape} is the thing that is \emph{preserved} by \texttt{fmap},
\emph{result} is the thing that is \emph{changed} by it. \texttt{fmap} cleanly
splits the two.

Interestingly, most introduction to Functors begin with describing functor
values as having a result and \texttt{fmap} as the thing that changes it, in
some way. Ironically, though it's a more common term, it's by far the more vague
and hard-to-intuit concept.

For something like \texttt{Maybe}, ``result'' is easy enough: it's the value
present if it exists. For parser-combinator \texttt{Parser}s too it's relatively
simple: the ``shape'' is the input consumed but the ``result'' is the Haskell
value you get as a result of the consumption. For \emph{optparse-applicative}
parser, it's the actual parsed command line arguments given by the user at
runtime. But sometimes it's more complicated: for the technical List functor,
the ``non-determinism'' functor, the ``shape'' is the number of options to
choose from and the order you get them in, and the ``result'' (to use precise
semantics) is the non-deterministic choice that you eventually pick or iterate
over.

So, the ``result'' can become a bit confusing to generalize. So, in my mind, I
usually reduce the definitions to:

\begin{itemize}
\tightlist
\item
  \emph{Shape}: the ``thing'' that \texttt{fmap} preserves: the \texttt{f} in
  \texttt{f\ a}
\item
  \emph{Result}: the ``thing'' that \texttt{fmap} changes: the \texttt{a} in
  \texttt{f\ a}
\end{itemize}

With this you could ``derive'' the Functor laws:

\begin{itemize}
\tightlist
\item
  \texttt{fmap\ id\ ==\ id}: \texttt{fmap} leaves the shape unchanged,
  \texttt{id} leaves the result unchanged. So entire thing must remain
  unchanged!
\item
  \texttt{fmap\ f\ .\ fmap\ g\ ==\ fmap\ (f\ .\ g)}. In both cases the shape
  remains unchanged, but one changes the result by f after g, and the other
  changes the result by \texttt{f\ .\ g}. They must be the same transformation!
\end{itemize}

All neat and clean, right? So, maybe the big misdirection is focusing too much
on the ``result'' when learning Functors, when we \emph{should} really be
focusing more on the ``shape'', or at least the two together.

Once you internalize ``\texttt{Functor} gives you shape-preservation'', this
helps you understand the value of the other common typeclass abstractions in
Haskell as well, and how they function based on how they manipulate ``shape''
and ``result''.

\section{Traversable}\label{traversable}

For example, what does the \texttt{Traversable} typeclass give us? Well, if
\texttt{Functor} gives us a way to map \emph{pure} functions and preserve shape,
then \texttt{Traversable} gives us a way to map \emph{effectful} functions and
preserve shape.

Whenever someone asks me about my favorite \texttt{Traversable} instance, I
always say it's the \texttt{Map\ k} traversable:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{traverse}\OtherTok{ ::} \DataTypeTok{Applicative}\NormalTok{ f }\OtherTok{=\textgreater{}}\NormalTok{ (a }\OtherTok{{-}\textgreater{}}\NormalTok{ f b) }\OtherTok{{-}\textgreater{}} \DataTypeTok{Map}\NormalTok{ k a }\OtherTok{{-}\textgreater{}}\NormalTok{ f (}\DataTypeTok{Map}\NormalTok{ k b)}
\end{Highlighting}
\end{Shaded}

Notice how it has no constraints on \texttt{k}? Amazing isn't it?
\texttt{Map\ k\ b} lets us map an \texttt{(a\ -\textgreater{}\ f\ b)} over the
values at each key in a map, and collects the results under the key the
\texttt{a} was originally under.

In essence, you can be assured that the result map has the \emph{same keys} as
the original map, perfectly preserving the ``shape'' of the map. The
\texttt{Map\ k} instance is the epitome of beautiful \texttt{Traversable}
instances. We can recognize this by identifying the ``shape'' that
\texttt{traverse} is forced to preserve.

\section{Applicative}\label{applicative}

What does the \texttt{Applicative} typeclass give us? It has \texttt{ap} and
\texttt{pure}, but
\href{https://hackage.haskell.org/package/base/docs/Control-Applicative.html\#t:Applicative}{its
laws} are infamously difficult to understand.

But, look at \texttt{liftA2\ (,)}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{liftA2 (,)}\OtherTok{ ::} \DataTypeTok{Applicative}\NormalTok{ f }\OtherTok{=\textgreater{}}\NormalTok{ f a }\OtherTok{{-}\textgreater{}}\NormalTok{ f b }\OtherTok{{-}\textgreater{}}\NormalTok{ f (a, b)}
\end{Highlighting}
\end{Shaded}

It lets us take ``two things'' and \emph{combine their shapes}. And, more
importantly, it combines the shapes \emph{without considering the results}.

\begin{itemize}
\tightlist
\item
  For \texttt{Writer\ w}, \texttt{\textless{}*\textgreater{}} lets us combine
  the two logged values using \texttt{mappend} while ignoring the actual
  \texttt{a}/\texttt{b} results.
\item
  For list, \texttt{\textless{}*\textgreater{}} (the cartesian product) lets us
  multiply the lengths of the input lists together. The \emph{length} of the new
  list ignores the actual \emph{contents} of the list.
\item
  For \texttt{State\ s}, \texttt{\textless{}*\textgreater{}} lets you
  \emph{compose} the \texttt{s\ -\textgreater{}\ s} state functions together,
  ignoring the \texttt{a}/\texttt{b}s
\item
  For \texttt{Parser}, \texttt{\textless{}*\textgreater{}} lets you sequence
  input consumption in a way that doesn't depend on the actual values you parse:
  it's ``context-free'' in a sense, aside from
  \href{https://byorgey.wordpress.com/2012/01/05/parsing-context-sensitive-languages-with-applicative/}{some
  caveats}.
\item
  For \emph{optparse-applicative}, \texttt{\textless{}*\textgreater{}} lets you
  combine your command line argument specs together, without depending on the
  actual values provided at runtime by the caller.
\end{itemize}

The key takeaway is that the ``final shape'' \emph{only depends} on the input
shapes, and not the results. You can know the length of
\texttt{\textless{}*\textgreater{}}-ing two lists together with only knowing the
length of the input lists, and you can also know the relative ordering of inputs
to outputs. Within the specific context of the semantics of \texttt{IO}, you can
know what ``effect'' \texttt{\textless{}*\textgreater{}}-ing two IO actions
would produce only knowing the effects of the input IO actions\footnote{That is,
  if we take the sum consideration of all input-output with the outside world,
  independent of what happens within the Haskell results, we can say the
  combination of effects is deterministic.}. You can know what command line
arguments \texttt{\textless{}*\textgreater{}}-ing two
\emph{optparse-applicative} parsers would have only knowing the command line
arguments in the input parsers. You can know what strings
\texttt{\textless{}*\textgreater{}}-ing two parser-combinator parsers would
consume or reject, based only on the consumption/rejection of the input parsers.
You can know the final log of \texttt{\textless{}*\textgreater{}}-ing two
\texttt{Writer\ w\ a}s together by only knowing the logs of the input writer
actions.

And hey\ldots some of these combinations feel ``monoidal'', don't they?

\begin{itemize}
\tightlist
\item
  \texttt{Writer\ w} sequences using \texttt{mappend}
\item
  List lengths sequence by multiplication
\item
  \texttt{State\ s} functions sequence by composition
\end{itemize}

You can also imagine ``no-op'' actions:

\begin{itemize}
\tightlist
\item
  \texttt{Writer\ w}'s no-op action would log \texttt{mempty}, the identity of
  \texttt{mappend}
\item
  List's no-op action would have a length 1, the identity of multiplication
\item
  \texttt{State\ s}'s no-op action would be \texttt{id}, the identity of
  function composition
\end{itemize}

That might sound familiar --- these are all \texttt{pure} from the
\texttt{Applicative} typeclass!

So, the Applicative typeclass laws aren't that mysterious at all. If you
understand the ``shape'' that a Functor induces, \texttt{Applicative} gives you
a \emph{monoid} on that shape! This is why \texttt{Applicative} is often called
the ``higher-kinded'' \texttt{Monoid}.

This intuition takes you pretty far, I believe. Look at the examples above where
we clearly identify specific \texttt{Applicative} instances with specific
\texttt{Monoid} instances (\texttt{Monoid\ w}, \texttt{Monoid\ (Product\ Int)},
\texttt{Monoid\ (Endo\ s)}).

Put in code:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} A part of list\textquotesingle{}s shape is its length and the monoid is (*, 1)}
\FunctionTok{length}\NormalTok{ (xs }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ ys) }\OperatorTok{==} \FunctionTok{length}\NormalTok{ xs }\OperatorTok{*} \FunctionTok{length}\NormalTok{ ys}
\FunctionTok{length}\NormalTok{ (}\FunctionTok{pure}\NormalTok{ r) }\OperatorTok{==} \DecValTok{1}

\CommentTok{{-}{-} Maybe\textquotesingle{}s shape is isJust and the monoid is (\&\&, True)}
\NormalTok{isJust (mx }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ my) }\OperatorTok{==}\NormalTok{ isJust mx }\OperatorTok{\&\&}\NormalTok{ isJust my}
\NormalTok{isJust (}\FunctionTok{pure}\NormalTok{ r) }\OtherTok{=} \DataTypeTok{True}

\CommentTok{{-}{-} State\textquotesingle{}s shape is execState and the monoid is (flip (.), id)}
\NormalTok{execState (sx }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ sy) }\OperatorTok{==}\NormalTok{ execState sy }\OperatorTok{.}\NormalTok{ execState sx}
\NormalTok{execState (}\FunctionTok{pure}\NormalTok{ r) }\OperatorTok{==} \FunctionTok{id}

\CommentTok{{-}{-} Writer\textquotesingle{}s shape is execWriter and the monoid is (\textless{}\textgreater{}, mempty)}
\NormalTok{execWriter (wx }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ wy) }\OperatorTok{==}\NormalTok{ execWriter wx }\OperatorTok{\textless{}\textgreater{}}\NormalTok{ execWriter wy}
\NormalTok{execWriter (}\FunctionTok{pure}\NormalTok{ r) }\OperatorTok{==} \FunctionTok{mempty}
\end{Highlighting}
\end{Shaded}

We can also extend this to non-standard \texttt{Applicative} instances: the
\texttt{ZipList} newtype wrapper gives us an \texttt{Applicative} instance for
lists where \texttt{\textless{}*\textgreater{}} is \texttt{zipWith}. These two
have the same \texttt{Functor} instances, so their ``shape'' (length) is the
same. And for both the normal \texttt{Applicative} and the \texttt{ZipList}
\texttt{Applicative}, you can know the length of the result based on the lengths
of the input, but \texttt{ZipList} combines shapes using the \texttt{Min}
monoid, instead of the \texttt{Product} monoid. And the identity of \texttt{Min}
is positive infinity, so \texttt{pure} for \texttt{ZipList} is an infinite list.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} A part of ZipList\textquotesingle{}s shape is length and its monoid is (min, infinity)}
\FunctionTok{length}\NormalTok{ (xs }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ ys) }\OperatorTok{==} \FunctionTok{length}\NormalTok{ xs }\OtherTok{\textasciigrave{}min\textasciigrave{}} \FunctionTok{length}\NormalTok{ ys}
\FunctionTok{length}\NormalTok{ (}\FunctionTok{pure}\NormalTok{ r) }\OperatorTok{==}\NormalTok{ infinity}
\end{Highlighting}
\end{Shaded}

The ``know-the-shape-without-knowing-the-results'' property is actually
leveraged by many libraries. It's how \emph{optparse-applicative} can give you
\texttt{-\/-help} output: the \emph{shape} of the optparse-applicative parser
(the command line arguments list) can be computed \emph{without knowing the
results} (the actual arguments themselves at runtime). You can list out what
arguments are expecting without ever getting any input from the user.

This is also leveraged by the
\emph{\href{https://hackage.haskell.org/package/async}{async}} library to give
us the \texttt{Concurrently} \texttt{Applicative} instance. Normally
\texttt{\textless{}*\textgreater{}} for IO gives us sequential combination of IO
effects. But, \texttt{\textless{}*\textgreater{}} for \texttt{Concurrently}
gives us \emph{parallel} combination of IO effects. We can launch all of the IO
effects in parallel at the same time because \emph{we know what the IO effects
are} before we actually have to execute them to get the results. If we needed to
know the results, this wouldn't be possible.

This also gives some insight into the
\href{https://hackage.haskell.org/package/transformers-0.6.1.2/docs/Control-Applicative-Backwards.html}{\texttt{Backwards}
Applicative wrapper} --- because the shape of the final does not depend on the
\emph{result} of either, we are free to combine the shapes in whatever order we
want. In the same way that every monoid gives rise to
\href{https://hackage.haskell.org/package/base-4.20.0.1/docs/Data-Monoid.html\#t:Dual}{a
``backwards'' monoid}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\OperatorTok{\textgreater{}} \StringTok{"hello"} \OperatorTok{\textless{}\textgreater{}} \StringTok{"world"}
\StringTok{"helloworld"}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ getDual }\OperatorTok{$} \DataTypeTok{Dual} \StringTok{"hello"} \OperatorTok{\textless{}\textgreater{}} \DataTypeTok{Dual} \StringTok{"world"}
\StringTok{"worldhello"}
\end{Highlighting}
\end{Shaded}

Every \texttt{Applicative} gives rise to a ``backwards'' \texttt{Applicative}
that does the shape ``mappending'' in reverse order:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\OperatorTok{\textgreater{}} \FunctionTok{putStrLn} \StringTok{"hello"} \OperatorTok{*\textgreater{}} \FunctionTok{putStrLn} \StringTok{"world"}
\NormalTok{hello}
\NormalTok{world}
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ forwards }\OperatorTok{$} \DataTypeTok{Backwards}\NormalTok{ (}\FunctionTok{putStrLn} \StringTok{"hello"}\NormalTok{) }\OperatorTok{*\textgreater{}} \DataTypeTok{Backwards}\NormalTok{ (}\FunctionTok{putStrLn} \StringTok{"world"}\NormalTok{)}
\NormalTok{world}
\NormalTok{hello}
\end{Highlighting}
\end{Shaded}

The monoidal nature of Applicative with regards to shapes and effects is the
heart of the original intent, and I've discussed this
\href{https://blog.jle.im/entry/const-applicative-and-monoids.html}{in earlier
blog posts}.

\section{Alternative}\label{alternative}

The main function of the
\emph{\href{https://hackage.haskell.org/package/base-4.20.0.1/docs/Control-Applicative.html\#g:2}{Alternative}}
typeclass is \texttt{\textless{}\textbar{}\textgreater{}}:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{(\textless{}|\textgreater{}) ::} \DataTypeTok{Alternative}\NormalTok{ f }\OtherTok{=\textgreater{}}\NormalTok{ f a }\OtherTok{{-}\textgreater{}}\NormalTok{ f a }\OtherTok{{-}\textgreater{}}\NormalTok{ f a}
\end{Highlighting}
\end{Shaded}

At first this might look a lot like \texttt{\textless{}*\textgreater{}} or
\texttt{liftA2\ (,)}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{liftA2 (,)}\OtherTok{ ::} \DataTypeTok{Applicative}\NormalTok{ f }\OtherTok{=\textgreater{}}\NormalTok{ f a }\OtherTok{{-}\textgreater{}}\NormalTok{ f b }\OtherTok{{-}\textgreater{}}\NormalTok{ f (a, b)}
\end{Highlighting}
\end{Shaded}

Both of them take two \texttt{f\ a} values and squish them into a single one.
Both of these are also monoidal on the shape, independent of the result. They
have a \emph{different} monoidal action on
\texttt{\textless{}\textbar{}\textgreater{}} than as
\texttt{\textless{}*\textgreater{}}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} A part of list\textquotesingle{}s shape is its length:}
\CommentTok{{-}{-} the Ap monoid is (*, 1), the Alt monoid is (+, 0)}
\FunctionTok{length}\NormalTok{ (xs }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ ys) }\OperatorTok{==} \FunctionTok{length}\NormalTok{ xs }\OperatorTok{*} \FunctionTok{length}\NormalTok{ ys}
\FunctionTok{length}\NormalTok{ (}\FunctionTok{pure}\NormalTok{ r) }\OperatorTok{==} \DecValTok{1}
\FunctionTok{length}\NormalTok{ (xs }\OperatorTok{\textless{}|\textgreater{}}\NormalTok{ ys) }\OperatorTok{==} \FunctionTok{length}\NormalTok{ xs }\OperatorTok{+} \FunctionTok{length}\NormalTok{ ys}
\FunctionTok{length}\NormalTok{ empty }\OperatorTok{==} \DecValTok{0}

\CommentTok{{-}{-} Maybe\textquotesingle{}s shape is isJust:}
\CommentTok{{-}{-} The Ap monoid is (\&\&, True), the Alt monoid is (||, False)}
\NormalTok{isJust (mx }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ my) }\OperatorTok{==}\NormalTok{ isJust mx }\OperatorTok{\&\&}\NormalTok{ isJust my}
\NormalTok{isJust (}\FunctionTok{pure}\NormalTok{ r) }\OtherTok{=} \DataTypeTok{True}
\NormalTok{isJust (mx }\OperatorTok{\textless{}|\textgreater{}}\NormalTok{ my) }\OperatorTok{==}\NormalTok{ isJust mx }\OperatorTok{||}\NormalTok{ isJust my}
\NormalTok{isJust empty }\OtherTok{=} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}

If we understand that functors have a ``shape'', \texttt{Applicative} implies
that the shapes are monoidal, and \texttt{Alternative} implies that the shapes
are a ``double-monoid''. The exact nature of how the two monoids relate to each
other, however, is not universally agreed upon. For many instances, however, it
does happen to form a \href{https://en.wikipedia.org/wiki/Semiring}{semiring},
where \texttt{empty} ``annihilates'' via
\texttt{empty\ \textless{}*\textgreater{}\ x\ ==\ empty}, and
\texttt{\textless{}*\textgreater{}} distributes over
\texttt{\textless{}\textbar{}\textgreater{}} like
\texttt{x\ \textless{}*\textgreater{}\ (y\ \textless{}\textbar{}\textgreater{}\ z)\ ==\ (x\ \textless{}*\textgreater{}\ y)\ \textless{}\textbar{}\textgreater{}\ (x\ \textless{}*\textgreater{}\ z)}.
But this is not universal.

However, what does \texttt{Alternative} bring to our shape/result dichotomy that
\texttt{Applicative} did not? Notice the subtle difference between the two:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{liftA2 (,)}\OtherTok{ ::} \DataTypeTok{Applicative}\NormalTok{ f }\OtherTok{=\textgreater{}}\NormalTok{ f a }\OtherTok{{-}\textgreater{}}\NormalTok{ f b }\OtherTok{{-}\textgreater{}}\NormalTok{ f (a, b)}
\OtherTok{(\textless{}|\textgreater{}) ::} \DataTypeTok{Alternative}\NormalTok{ f }\OtherTok{=\textgreater{}}\NormalTok{ f a }\OtherTok{{-}\textgreater{}}\NormalTok{ f a }\OtherTok{{-}\textgreater{}}\NormalTok{ f a}
\end{Highlighting}
\end{Shaded}

For \texttt{Applicative}, the ``result'' comes from the results of both inputs.
For \texttt{Alternative}, the ``result'' could come from one or the other input.
So, this introduces a fundamental data dependency for the \emph{results}:

\begin{itemize}
\tightlist
\item
  Applicative: Shapes merge monoidally independent of the results, but to get
  the result of the final, you need to produce the results of both of the two
  inputs in the general case.
\item
  Alternative: Shapes merge monoidally independent of the results, but to get
  the result of the final, you need the results of one or the other input in the
  general case.
\end{itemize}

This also implies that choice of combination method for shapes in
\texttt{Applicative} vs \texttt{Alternative} aren't arbitrary: the former has to
be ``conjoint'' in a sense, and the latter has to be ``disjoint''.

See again that clearly separating the shape and the result gives us the
vocabulary to say precisely what the different data dependencies are.

\section{Monad}\label{monad}

Understanding shapes and results also help us appreciate more the sheer
\emph{power} that Monad gives us. Look at
\texttt{\textgreater{}\textgreater{}=}:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{(\textgreater{}\textgreater{}=) ::} \DataTypeTok{Monad}\NormalTok{ m }\OtherTok{=\textgreater{}}\NormalTok{ m a }\OtherTok{{-}\textgreater{}}\NormalTok{ (a }\OtherTok{{-}\textgreater{}}\NormalTok{ m b) }\OtherTok{{-}\textgreater{}}\NormalTok{ m b}
\end{Highlighting}
\end{Shaded}

Using \texttt{\textgreater{}\textgreater{}=} means that the shape of the final
action is allowed to \emph{depend on the result} of the first action! We are no
longer in the Applicative/Alternative world where shape only depends on shape.

Now we can write things like:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{greet }\OtherTok{=} \KeywordTok{do}
  \FunctionTok{putStrLn} \StringTok{"What is your name?"}
\NormalTok{  n }\OtherTok{\textless{}{-}} \FunctionTok{getLine}
  \FunctionTok{putStrLn}\NormalTok{ (}\StringTok{"Hello, "} \OperatorTok{++}\NormalTok{ n }\OperatorTok{++} \StringTok{"!"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Remember that for ``IO'', the shape is the IO effects (In this case, what
exactly gets sent to the terminal) and the ``result'' is the haskell value
computed from the execution of that IO effect. In our case, the \emph{action} of
the result (what values are printed) depends on the \emph{result} of of the
intermediate actions (the \texttt{getLine}). You can no longer know in advance
what action the program will have without actually running it and getting the
results.

The same thing happens when you start sequencing parser-combinator parsers: you
can't know what counts as a valid parse or how much a parser will consume until
you actually start parsing and getting your intermediate parse results.

\texttt{Monad} is also what makes \texttt{guard} and co. useful. Consider the
purely Applicative:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{evenProducts ::}\NormalTok{ [}\DataTypeTok{Int}\NormalTok{] }\OtherTok{{-}\textgreater{}}\NormalTok{ [}\DataTypeTok{Int}\NormalTok{] }\OtherTok{{-}\textgreater{}}\NormalTok{ [}\DataTypeTok{Bool}\NormalTok{]}
\NormalTok{evenProducts xs ys }\OtherTok{=}\NormalTok{ (\textbackslash{}x y }\OtherTok{{-}\textgreater{}} \FunctionTok{even}\NormalTok{ (x }\OperatorTok{*}\NormalTok{ y)) }\OperatorTok{\textless{}$\textgreater{}}\NormalTok{ xs }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ ys}
\end{Highlighting}
\end{Shaded}

If you passed in a list of 100 items and a list of 200 items, you can know that
the result has 100 * 200 = 20000 items, without actually knowing any of the
items in the list.

But, consider an alternative formulation where we are allowed to use Monad
operations:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{evenProducts ::}\NormalTok{ [}\DataTypeTok{Int}\NormalTok{] }\OtherTok{{-}\textgreater{}}\NormalTok{ [}\DataTypeTok{Int}\NormalTok{] }\OtherTok{{-}\textgreater{}}\NormalTok{ [(}\DataTypeTok{Int}\NormalTok{, }\DataTypeTok{Int}\NormalTok{)]}
\NormalTok{evenProducts xs ys }\OtherTok{=} \KeywordTok{do}
\NormalTok{  x }\OtherTok{\textless{}{-}}\NormalTok{ xs}
\NormalTok{  y }\OtherTok{\textless{}{-}}\NormalTok{ ys}
\NormalTok{  guard (}\FunctionTok{even}\NormalTok{ (x }\OperatorTok{*}\NormalTok{ y))}
  \FunctionTok{pure}\NormalTok{ (x, y)}
\end{Highlighting}
\end{Shaded}

Now, \emph{even if you knew} the lengths of the input lists, you can \emph{not}
know the length of the output list without actually knowing what's inside your
lists. You need to actually start ``sampling''.

That's why there is no \texttt{Monad} instance for \texttt{Backwards} or
\emph{optparse-applicative} parsers. For \texttt{Backwards} doesn't work because
we've now introduced an asymmetry (the \texttt{m\ b} depends on the \texttt{a}
of the \texttt{m\ a}) that can't be reversed. For \emph{optparse-applicative},
it's because we want to be able to inspect the shape without knowing the results
at runtime (so we can show a useful \texttt{-\/-help} without getting any actual
arguments): but, with \texttt{Monad}, we can't know the shape without knowing
the results!

In a way, Monad simply ``is'' the way to combine \texttt{Functor} shapes
together where the final shape is allowed to depend on the results. Hah, I
tricked you into reading a monad tutorial!

\section{Free Structures}\label{free-structures}

I definitely write
\href{https://blog.jle.im/entry/functor-combinatorpedia.html}{way too much about
free structures} on this blog. But this ``shapeful'' way of thinking also gives
rise to why free structures are so compelling and interesting to work with in
Haskell.

Before, we were describing shapes of Functors and Applicatives and Monads that
already existed. We had \emph{this} \texttt{Functor}, what was \emph{its} shape?

However, what if we had a shape that we had in mind, and wanted to \emph{create}
an \texttt{Applicative} or \texttt{Monad} that manipulated that shape?

For example, let's roll our own version of \emph{optparse-applicative} that only
supported \texttt{-\/-myflag\ somestring} options. We could say that the
``shape'' is the list of supported option and parsers. So a single element of
this shape would be the specification of a single option:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Option}\NormalTok{ a }\OtherTok{=} \DataTypeTok{Option}\NormalTok{ \{}\OtherTok{ optionName ::} \DataTypeTok{String}\NormalTok{,}\OtherTok{ optionParse ::} \DataTypeTok{String} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a \}}
  \KeywordTok{deriving} \DataTypeTok{Functor}
\end{Highlighting}
\end{Shaded}

The ``shape'' here is the name and also what values it would parse, essentially.
\texttt{fmap} won't affect the name of the option and won't affect what would
succeed or fail.

Now, to create a full-fledged multi-argument parser, we can use
\href{https://hackage.haskell.org/package/free/docs/Control-Applicative-Free.html}{\texttt{Ap}
from the \emph{free} library}:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{type} \DataTypeTok{Parser} \OtherTok{=} \DataTypeTok{Ap} \DataTypeTok{Option}
\end{Highlighting}
\end{Shaded}

We specified the shape we wanted, now we get the \texttt{Applicative} of that
shape for free! We can now combine our shapes monoidally using the
\texttt{\textless{}*\textgreater{}} instance, and then use \texttt{runAp\_} to
inspect it:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Args} \OtherTok{=} \DataTypeTok{Args}\NormalTok{ \{}\OtherTok{ myStringOpt ::} \DataTypeTok{String}\NormalTok{,}\OtherTok{ myIntOpt ::} \DataTypeTok{Int}\NormalTok{ \}}

\OtherTok{parseTwo ::} \DataTypeTok{Parser}\NormalTok{ args}
\NormalTok{parseTwo }\OtherTok{=} \DataTypeTok{Args} \OperatorTok{\textless{}$\textgreater{}}\NormalTok{ liftAp stringOpt }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ liftAp intOpt}
  \KeywordTok{where}
\NormalTok{    stringOpt }\OtherTok{=} \DataTypeTok{Option} \StringTok{"string{-}opt"} \DataTypeTok{Just}
\NormalTok{    intOpt }\OtherTok{=} \DataTypeTok{Option} \StringTok{"int{-}opt"}\NormalTok{ readMaybe}

\OtherTok{getAllOptions ::} \DataTypeTok{Parser}\NormalTok{ a }\OtherTok{{-}\textgreater{}}\NormalTok{ [}\DataTypeTok{String}\NormalTok{]}
\NormalTok{getAllOptions }\OtherTok{=}\NormalTok{ runAp\_ (\textbackslash{}o }\OtherTok{{-}\textgreater{}}\NormalTok{ [optionName o])}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\OperatorTok{\textgreater{}}\NormalTok{ getAllOptions parseTwo}
\NormalTok{[}\StringTok{"string{-}opt"}\NormalTok{, }\StringTok{"int{-}opt"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

Remember that \texttt{Applicative} is like a ``monoid'' for shapes, so
\texttt{Ap} gives you a free ``monoid'' on your custom shape: you can now create
list-like ``sequences'' of your shape that merge via concatenation through
\texttt{\textless{}*\textgreater{}}. You can also know that \texttt{fmap} on
\texttt{Ap\ Option} will not add or remove options: it'll leave the actual
options unchanged. It'll also not affect what options would fail or succeed to
parse.

You could also write a parser combinator library this way too! Remember that the
``shape'' of a parser combinator \texttt{Parser} is the string that it consumes
or rejects. The single element might be a parser that consumes and rejects a
single \texttt{Char}:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{newtype} \DataTypeTok{Single}\NormalTok{ a }\OtherTok{=} \DataTypeTok{Single}\NormalTok{ \{}\OtherTok{ satisfies ::} \DataTypeTok{Char} \OtherTok{{-}\textgreater{}} \DataTypeTok{Maybe}\NormalTok{ a \}}
  \KeywordTok{deriving} \DataTypeTok{Functor}
\end{Highlighting}
\end{Shaded}

The ``shape'' is whether or not it consumes or rejects a char. Notice that
\texttt{fmap} for this cannot \emph{change} whether or not a char is rejected or
accepted: it can only change the Haskell result \texttt{a} value. \texttt{fmap}
can't flip the \texttt{Maybe} into a \texttt{Just} or \texttt{Nothing}.

Now we can create a full monadic parser combinator library by using
\href{https://hackage.haskell.org/package/free/docs/Control-Monad-Free.html}{\texttt{Free}
from the \emph{free} library}:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{type} \DataTypeTok{Parser} \OtherTok{=} \DataTypeTok{Free} \DataTypeTok{Single}
\end{Highlighting}
\end{Shaded}

Again, we specified the shape we wanted, and now we have a Monad for that shape!
For more information on using this, I've written
\href{https://blog.jle.im/entry/free-alternative-regexp.html}{a blog post in the
past}. \texttt{Ap} gives you a free ``monoid'' on your shapes, but in a way
\texttt{Free} gives you a ``tree'' for your shapes, where the sequence of shapes
depends on which way you go down their results. And, again, \texttt{fmap} won't
ever change what would or would not be parsed.

How do we know what free structure to pick? Well, we ask questions about what we
want to be able to do with our shape. If we want to inspect the shape without
knowing the results, we'd use the free Applicative or free Alternative. As
discussed earlier, using the free Applicative means that our final result must
require producing all of the input results, but using the free Alternative means
it doesn't. If we wanted to allow the shape to depend on the results (like for a
context-sensitive parser), we'd use the free Monad. Understanding the concept of
the ``shape'' makes this choice very intuitive.

\section{The Shape of You}\label{the-shape-of-you}

Next time you encounter a new Functor, I hope these insights can be useful. Ask
yourself, what is \texttt{fmap} preserving? What is \texttt{fmap} changing? And
from there, its secrets will unfold before you.
\href{https://en.wikipedia.org/wiki/Noether\%27s_theorem}{Emmy Noether} would be
proud.

\section{Special Thanks}\label{special-thanks}

I am very humbled to be supported by an amazing community, who make it possible
for me to devote time to researching and writing these posts. Very special
thanks to my supporter at the ``Amazing'' level on
\href{https://www.patreon.com/justinle/overview}{patreon}, Josh Vera! :)

\section{Signoff}\label{signoff}

Hi, thanks for reading! You can reach me via email at
\href{mailto:justin@jle.im}{\nolinkurl{justin@jle.im}}, or at twitter at
\href{https://twitter.com/mstk}{@mstk}! This post and all others are published
under the \href{https://creativecommons.org/licenses/by-nc-nd/3.0/}{CC-BY-NC-ND
3.0} license. Corrections and edits via pull request are welcome and encouraged
at \href{https://github.com/mstksg/inCode}{the source repository}.

If you feel inclined, or this post was particularly helpful for you, why not
consider \href{https://www.patreon.com/justinle/overview}{supporting me on
Patreon}, or a \href{bitcoin:3D7rmAYgbDnp4gp4rf22THsGt74fNucPDU}{BTC donation}?
:)

\end{document}
