\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{\textbf{#1}}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={I nixified my blog},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
% Make links footnotes instead of hotlinks:
\renewcommand{\href}[2]{#2\footnote{\url{#1}}}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\title{I nixified my blog}

\begin{document}
\maketitle

\% Justin Le \% January 1, 2023

\emph{Originally posted on
\textbf{\href{https://blog.jle.im/entry/i-nixified-my-blog.html}{in Code}}.}

Happy new year! It's been a while (almost three years) since my last post. But
it's a new year, a new me -- and I just recently started working at
\href{https://github.com/anduril}{a new job} writing Haskell where everything is
built and deployed using \href{https://nixos.org/}{nix}. Aside from the bliss
that comes from being able to write Haskell as a day job, I also enjoyed the
push to finally dive into the nix ecosystem for my general personal computing,
which many of my friends have been subtly (or not so subtly) pushing me to look
into for a long time. And I have to say, I'm hooked! I get a lot of similar joy
using nix to organize my projects as I did when I was first learning Haskell. My
path to re-organizing all of my personal projects to using nix has lead me back
to one of my more ``longer-running'' pieces of legacy code -- my 10 year old
blog. So, this is a post from a naive new nix user on how I converted my blog
deployment and building from a manual multi-stage build process into an
automatic nix-with-cachix deploy -- and some future things I would be hoping to
investigate!

In this post I'll also be explaining a bit of \emph{nix}, so hopefully it's
accessible if you are curious like I was too. However, it's \emph{not} a
tutorial --- instead, it's a high-level overview of the concepts that I put
together to achieve the goal.

\section{How Did We Get Here}\label{how-did-we-get-here}

The \href{https://blog.jle.im/entries/category/@meta.html}{development history}
of my blog (which just turned 10 a few months ago, actually) has gone through a
few stages: (which you might be able to track
\href{https://github.com/mstksg/inCode}{on github})

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A simple \emph{\href{https://hackage.haskell.org/package/scotty}{scotty}}
  server on Heroku backed by a cloud-hosted postgresql database. My first ever
  non-trivial Haskell project!
\item
  Using \emph{\href{https://hackage.haskell.org/package/fay}{fay}} to compile to
  javascript for interactivity and post-processing in blog posts.
\item
  Converting \emph{fay} to \emph{ghcjs}, to write javascript in Haskell.
\item
  Switching from ``classic cabal'' to
  \emph{\href{https://docs.haskellstack.org/en/stable/}{stack}} for development
  and building.
\item
  Using \emph{\href{https://shakebuild.com/}{shake}} for build/deploy scripts.
\item
  Converting \emph{ghcjs} to
  \emph{\href{https://www.purescript.org/}{purescript}}, because I didn't want
  to bundle 1.5 MB of Haskell Runtime just to generate a table of contents.
\item
  The biggest change: moving from \emph{scotty} to
  \emph{\href{https://jaspervdj.be/hakyll/}{hakyll}} (a static site generator),
  from a simple HTTP server to a static site generation.
\item
  Deploying to github pages within a build script, using a custom haskell tool I
  had written a long time ago. I honestly don't remember where the site was
  hosted before github pages, but I remember having to configure my own CDN and
  https for a my hosted static website and at some point I wondered ``why?''.
\item
  Moving from yaml for configuration to
  \emph{\href{https://dhall-lang.org/}{dhall}}.
\end{enumerate}

As you can see, I sort of use my blog as a playground to apply new ideas I'm
learning build familiarity with them, and I have some fun refactoring every
couple of years because Haskell is so fun to refactor.

However, the current state of things is kind of a monster to build. First, you
have to have purescript \emph{and} npm \emph{and} bower (for general javascript
deployment) installed. Then the build script (the Shakefile) compiles javascript
and purescript into the working directory, then calls hakyll which generates the
actual site files. It's a really finicky system that involves having the correct
javascript ecosystem dependencies already installed -- and also uses a
deprecated system for building purescript. At least we had \emph{stack} to
manage having the correct ghc version.

Looking at this in 2023, there are a few things that had to change:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The purescript ecosystem has moved onto
  \emph{\href{https://github.com/purescript/spago}{spago}} for dependency
  management.
\item
  I have been moving most of my projects from stack back to pure cabal using the
  new modern tooling
\item
  I have moved dev environments a few times and every time it is a huge pain to
  remember the exact set of things that need to be installed for all stages to
  build and work correctly. I'd like to move from an ad-hoc build script
  dependent on the environment into a single tool that manages everything.
\item
  I wanted to get automatic deployments to github pages running instead of
  having to manually deploy.
\end{enumerate}

Some of these don't necessarily require nix, but I figured that this would be a
good opportunity to check everything off and seeing what nix had to offer.

\section{The Game Plan}\label{the-game-plan}

Going into this, the general workflow would be:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Create a \emph{nix derivation} for all of the purescript scripts, that
  generates a single javascript file each.
\item
  Create a nix derivation for the \emph{hakyll} static site generator binary
\item
  Create a nix derivation for the full static site itself: it will assemble all
  of the compiled purescript files from step 1, the source files (markdown,
  static images, etc.) from the repo, and then run the hakyll binary on it.
\item
  Create a github actions workflow to automatically build the full static site
  generation workflow.
\item
  Use \emph{\href{https://www.cachix.org/}{cachix}} to ensure that all
  derivations where the dependencies don't change will not need to be re-built.
\item
  If step 4 completes, automatically push the derivation's outputs to the
  \emph{gh-pages} branch.
\item
  Establish \emph{local development environments} that nix can set up for us:

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    Haskell development -- with haskell language server, cabal for intermediate
    building, etc.
  \item
    Purescript development -- I've structured each interactive blog post's
    purescript source as its own spago project, so we actually need to generate
    a separate development environment (with purescript language server, etc.)
    for each project. I don't expect to be working on more than one at a given
    time.
  \item
    \emph{Writing} -- with the hakyll binary available for intermediate caching
    for fast builds and updates and hakyll's local preview server, and the
    \emph{purescript} compiler available for fast compilation for the scripts
    driving interactive blog posts.
  \end{enumerate}
\end{enumerate}

One cool thing about cachix is that you get build caching for free for any
derivation you hand-write across all of your environments. You don't need to
register your ``package'' to any sort of registry or set up custom cloud caching
in every situation.

\subsection{What is a Derivation?}\label{what-is-a-derivation}

The main character in all of these steps is the \emph{nix derivation}. In my
mental model, a nix derivation is a recipe for building ``something'' --
sometimes, it's a binary, and in our case it's often just going to be a
collection of files. A derivation contains a list of pinned reproducible
dependency derivations --- so it's all reproducible down the whole chain. In our
case:

\begin{itemize}
\tightlist
\item
  The final derivation (the static site's html files and contents)
  \emph{depends} on the hakyll binary derivation and the compiled purescript
  derivation
\item
  The hakyll binary derivation depends on a pinned ghc, cabal, and a derivation
  for every haskell library dependency.
\item
  The compiled purescript derivation depends on the purescript compiler (purs),
  as well as all purescript library dependencies
\end{itemize}

When we say ``depends'', it just means that uses what the derivation outputs in
one way or another. The final static site derivation uses the \emph{binary}
outputted by the hakyll binary derivation. It also uses the \emph{javascript
files} outputted by the compiled purescript derivation. And its \emph{output} is
just the static pages and contents of the site.

One major leap for me was realizing that derivations just created output files.
In my mind, before this, I thought that a derivation was something
``executable'' --- when you first get into nix and nixos, ``derivation'' and
``executable'' can sometimes appear synonymous. To install a new executable in
nixos (like \emph{vim} or any other linux program), you installed its
derivation. However, this was a misconception. It's just a \emph{convention}
that if your derivation is meant to provide an executable, you put the binary
file in the \texttt{./bin/} subdirectory of your output files. In our case, our
output files are sometimes binaries (which we put in \texttt{./bin/}), but
they're also sometimes static files, which we put in \texttt{./dist/} or
\texttt{./js/} or wherever we want. It's just a ``convention'' that the
\texttt{./bin} subdirectory of your output contains a binary that the caller
would want in their PATH, and so nix automatically puts the \texttt{./bin} of
any dependency derivation in the PATH/execution environment for convenience. At
least, that's my mental model and how I understand it --- I am definitely new to
this, so if I'm incorrect at any point, feel free to leave a comment.

\subsection{What is a development
environment?}\label{what-is-a-development-environment}

Very important to this system is the local development environment that lets us
quickly re-build the site (and also lets us use hakyll's built-in preview server
features). I have a little less refined of a mental model for these than I do
for the derivation --- but apparently it's just a derivation that sets up a
shell with everything we need in the PATH? In any case, we need to set one up
here to drop us in a Haskell development environment to get all of our favorite
Haskell development tools -- like Haskell Language Server and cabal -- and a
snapshot of all of our Haskell dependencies. And then we need one that just
gives us the hakyll binary, which has interactive development features like
directory watching, local servers, and incremental builds/cache invalidation.

\subsection{Prior Art}\label{prior-art}

While researching this further, I noticed that this approach is similar to that
of
\emph{\href{https://github.com/rpearce/hakyll-nix-template}{hakyll-nix-template}}:
have a derivation for the hakyll binary, then a derivation for the final
website, then push to gh-pages with cachix to ensure caching haskell
dependencies. Because of this, I was able to use a lot of code from that
repository to help figure things out for my own path. So, a special thanks to
\href{https://github.com/rpearce}{Robert Pearce}!

\subsection{Shortcomings: Static Artifact
Caching}\label{shortcomings-static-artifact-caching}

There is one shortcoming to this approach that I consider very significant --
using nix with \emph{cachix} ensures that we don't redo work at the derivation
level. So, if the purescript files don't change, we won't re-build the compiled
javascript. If the haskell files don't change, we won't re-build the hakyll
binary. But for a static site, we also have to build the site based on the input
files --- mostly, markdown files. For my blog, this is not trivial -- each
markdown post gets compiled and outputted into many different formats (latex,
pdf, html multiple times), and actually gets re-combined with re-usable code
snippets from different folders. In addition, we build a page for each of the
many tags, a paginated homepage, a granular history page, and an rss feed. On a
whole, a build from scratch takes about five minutes. This is a significant
downgrade --- in the past, I could build and deploy almost instantaneously
(under a minute).

I'm not sure if this can be optimized on the Haskell side, but it's still kind
of wasteful in principle, because \emph{hakyll} is smart enough to only re-build
what is needed, if it's run in its normal mode of operation. If you update only
a single post, hakyll will only update the pages that are effected --- the home
page, the rss feed, and any tag pages it appears on, etc. It knows not to
re-generate any single blog post pages or pdf files for different blog posts
that aren't affected by the new update.

If you're using github actions in the normal way, you might be able to take
advantage of this with github actions caches? It's unclear to me how change
detection interacts with this. However, in the ``nix style'', if we really want
to view the final generated site as a \emph{nix derivation} (and not just, ie, a
build script), it's a little more complicated to make this fit into the
conceptual model and make sure cache invalidation etc. works correctly.

What would be really ideal is if these units of cached data (individual web
pages, etc.) could be managed and invalidated by nix itself --- maybe each blog
post's outputs (the html, pdf, post-processed markdown, latex) could be a
derivation, and so these derivations won't have to be re-built if the blog post
is not changed. There seems to be a few options in terms of ``purely nix''
static site generators that could do this I think, but that's not really an
option to me because the blog at this point is pretty huge, logic-wise, and also
I would rather write the blog in Haskell in the end still.

Maybe in the future I could figure out how to modify hakyll to use the nix cache
for its caching and invalidation, instead of its own bespoke method? I wonder if
that's even a viable option. If anyone has any insight, this newbie would really
appreciate hearing!

In any case, for now, we can still have fast local development building (more on
that later), so it isn't super painful to actually write test. However, the
deployment lag is still kind of annoying --- and the ``principal'' of it (the
wasted work) does still bother me.

\section{Assembling the pieces}\label{assembling-the-pieces}

Again, this isn't a tutorial --- here, I'm going to explain to explain the
components I put together at a high level.

We're going to structure this in terms of a nix flake, which I'm using to:

\begin{itemize}
\tightlist
\item
  Pin dependencies and package sets
\item
  Describe buildable derivations and the development environments in a single
  file, which we can run with \texttt{nix\ build} and \texttt{nix\ develop}
  commands.
\end{itemize}

For reference, the ``final'' nix flake described in this post
\href{https://github.com/mstksg/inCode/blob/bf444098b8773b1b47c0b2e66ca1682cdc5f674d/flake.nix}{can
be found on github}.

\subsection{Haskell}\label{haskell}

For Haskell, I ended up using
\emph{\href{https://input-output-hk.github.io/haskell.nix/\#haskellnix}{haskell.nix}},
which is what I also what we use at work. It does this neat thing where it
parses a cabal file and uses \emph{that} to generate the derivation needed ---
which includes the dependency derivations for all haskell library derivations.
In this way, it can cache all external dependencies. It also lets us solve for
which version of each dependency we need. And, by freezing it with flake lock,
it also allows us to freeze all of the versions of each dependency in a way that
works. There might be other tools that allow this, but I went with haskell.nix
because I'm already familiar with it. It also has a nice development environment
offered that contains haskell language server with all of the right package
versions and artifacts.

Armed with this, we can easily generate a derivation for the hakyll binary, by
passing the cabal file to the \texttt{haskell-nix.project\textquotesingle{}}
function provided by haskell.nix. Now step 1 is complete!

Note we also get our development environment as well -- you get it with the
\texttt{devShell} property!

\subsection{Purescript}\label{purescript}

For Purescript, I ended up using
\href{https://github.com/purifix/purifix}{purifix}, which does a similar thing
with \emph{haskell.nix} but for the \emph{spago.yaml} file, which is how modern
purescript projects declare their dependencies. Purifix also has this mode of
operation where it auto-detects if you're working in a monorepo-style project
with multiple binaries, and gives you a \emph{separate derivation} for each
binary (or, compiled .js file)! And also a separate development environment for
each binary, too. This works well for me because I structure the javascript for
each interactive blog post as its own project. So, I can export an entire
``project'' (blog post file) to a bundled file easily, and also lead development
environments for specific blog posts as well.

For each project, it provides a \texttt{bundle-app} derivation that is literally
just the generated bundled-up single javascript file. This is exactly the
derivation we need for the final part where we pull everything together.

However, it also gives us an easy way to access all of the derivations of all of
the dependencies (the \texttt{globs} property it adds onto its derivations) ---
which is very useful for our final integrated development environment which
needs to do incremental builds quickly.

A note -- finding all of these useful properties and derivations (outside of the
main ones) does take a bit of trial and error -- digging through
\texttt{nix\ repl} to explore the derivation contents, and the source code of
these projects.

\subsection{Static Site derivation}\label{static-site-derivation}

Putting it all together is the final derivation to generate the actual site:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{web = pkgs.stdenv.mkDerivation }\OperatorTok{\{}
  \VariableTok{name} \OperatorTok{=} \StringTok{"inCode"}\OperatorTok{;}
  \VariableTok{buildInputs} \OperatorTok{=} \OperatorTok{[}\NormalTok{ inCode.haskell }\OperatorTok{]} \OperatorTok{++}\NormalTok{ lib.attrValues inCode.purescript}\OperatorTok{;}
  \VariableTok{srcs} \OperatorTok{=} \OperatorTok{[}
    \SpecialStringTok{./code{-}samples}
    \SpecialStringTok{./config}
    \SpecialStringTok{./copy}
    \SpecialStringTok{./css}
    \SpecialStringTok{./js}
    \SpecialStringTok{./latex}
    \SpecialStringTok{./scss}
    \SpecialStringTok{./static}
  \OperatorTok{];}
  \VariableTok{unpackPhase} \OperatorTok{=} \StringTok{\textquotesingle{}\textquotesingle{}}
\StringTok{    for srcFile in $srcs; do}
\StringTok{      cp {-}a $srcFile/. $(stripHash $srcFile)}
\StringTok{    done}

\StringTok{    mkdir \_purescript}
\StringTok{    }\SpecialCharTok{$\{}
\NormalTok{      lib.concatStringsSep }\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"} \OperatorTok{(}\NormalTok{lib.mapAttrsToList}
          \OperatorTok{(}\VariableTok{name}\OperatorTok{:} \VariableTok{value}\OperatorTok{:} \StringTok{\textquotesingle{}\textquotesingle{}cp }\SpecialCharTok{$\{}\NormalTok{value.bundle}\OperatorTok{{-}}\NormalTok{app}\SpecialCharTok{\}}\StringTok{ \_purescript/}\SpecialCharTok{$\{}\NormalTok{name}\SpecialCharTok{\}}\StringTok{.js\textquotesingle{}\textquotesingle{}}\OperatorTok{)}
\NormalTok{          inCode.purescript}
        \OperatorTok{)}
     \SpecialCharTok{\}}
\StringTok{  \textquotesingle{}\textquotesingle{}}\OperatorTok{;}
  \VariableTok{buildPhase} \OperatorTok{=} \StringTok{\textquotesingle{}\textquotesingle{}}
\StringTok{    }\SpecialCharTok{$\{}\NormalTok{inCode.haskell}\SpecialCharTok{\}}\StringTok{/bin/inCode{-}build build {-}{-}verbose}
\StringTok{  \textquotesingle{}\textquotesingle{}}\OperatorTok{;}
  \VariableTok{installPhase} \OperatorTok{=} \StringTok{\textquotesingle{}\textquotesingle{}}
\StringTok{    mkdir {-}p "$out/dist"}
\StringTok{    cp {-}a \_site/. "$out/dist"}
\StringTok{  \textquotesingle{}\textquotesingle{}}\OperatorTok{;}
\OperatorTok{\}}\NormalTok{;}
\end{Highlighting}
\end{Shaded}

This is the only ``hand-built'' derivation. Hopefully it's legible enough:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The \texttt{buildInputs} attribute tells us what derivations give us the
  binaries/generated files in scope for us to do our job.

  \begin{itemize}
  \tightlist
  \item
    \texttt{inCode.haskell} is the derivation that gives us the hakyll binary,
    from running \emph{haskell.nix} on the haskell hakyll project.
  \item
    \texttt{lib.attrValues\ inCode.purescript} gives us each of the derivations
    of all of the purescript projects, one for each spago sub-project (and each
    blog post/page that uses purescript). In this case, it gives us the actual
    compiled javascript bundle that the site generator expects.
  \end{itemize}
\item
  \texttt{srcs} is the actual source files that the static site generator uses
  --- in this case, the code samples, configuration files, markdown files for
  the actual blog posts in \texttt{./copy}, the static files, etc.
\item
  \texttt{unpackPhase} is the shell script to get things ready for the hakyll
  site generator to run:

  \begin{itemize}
  \tightlist
  \item
    First, copy all of the source files into the temporary build directory
  \item
    Then, copy all of the compiled javascript binaries (from the
    \texttt{bundle-app} property of each derivation) into the
    \texttt{\_purescript} folder in the temporary build directory, where the
    static site generator expects to find it.
  \end{itemize}
\item
  \texttt{buildPhase} is the shell script to actually run the static site
  generator on the work directory that we carefully unpacked. Note that we refer
  to the binary using the derivation variable, \texttt{\$\{inCode.haskell\}} and
  knowing that it is found under the \texttt{bin} folder, as convention states.
\item
  The \texttt{installPhase} copies what the site generated (in the
  \texttt{\_site} directory of the temporary work folder) into \texttt{\$out},
  which is the place we put the ``results''.
\end{enumerate}

\subsection{Github Action}\label{github-action}

The final github action is pretty standard --- it pulls together the actions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{cachix/install-nix-action}: Install nix
\item
  \texttt{cachix/cachix-action}: Use the given cachix cache and upload the
  artifacts to it after building
\item
  Actually run \texttt{nix\ build} on the static site derivation to generate the
  static files. In the process, it will either re-build (if anything changes)
  the dependency haskell and purescfript derivations, or pull it straight from
  cachix if nothing changed.
\item
  \texttt{crazy-max/ghaction-github-pages} to push the files to the
  \texttt{gh-pages} branch
\end{enumerate}

This was taken pretty much verbatim from \emph{hakyll-nix-template}.

\subsection{Development Environment for
Writing}\label{development-environment-for-writing}

Now, because we don't have page-level caching for nix (and hakyll manages its
own caching), we have to bite the bullet and create a customized development
environment where we can locally deploy hakyll's cached incremental builds.
Again, this will be a hand-written environment, so let's think about what we
want. We need ``temporary'' directories that hakyll will use to (1) find the
source files, (2) store its cache, and (3) output the static site. Because of
this, if we make any changes to blog posts while we are in our development
environment, hakyll will not do a full re-build --- it'll still have its
development-environment-scoped cache. We can also use hakyll's built in local
preview server and filesystem watch-and-rebuild features. We also want to do
incremental builds for purescript development as well, and we need to make sure
that this temporary environment also has an incremental build cache area for
\texttt{purs}. While we \emph{could} do this without any temporary directories
by just using the working directory for this, it is kind of nice to be able to
have this all take place in a temporary folder that will go away after we exit
the development environment.

Unfortunately, the hakyll binary isn't aware of \texttt{purs} and purescript ---
it just receives the compiled bundles. So, we need a way for the user to run a
command to re-build the purescript dependencies if they have changed it. There
are a few ways we can do this (maybe have nix initiate a ``watcher'' that
automatically incrementally recompiles, or wrap our hakyll binary in a wrapper
shell script that triggers a re-build). However, the simplest way I could think
of (that also does the job) is for the development environment to provide a
command/shell script that just runs \texttt{purs} with the right arguments.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rebuild}\OperatorTok{{-}}\NormalTok{js =}
  \KeywordTok{let}
    \VariableTok{buildSingleDep} \OperatorTok{=} \VariableTok{name}\OperatorTok{:} \VariableTok{value}\OperatorTok{:}
      \KeywordTok{let}
        \VariableTok{srcGlob} \OperatorTok{=} \StringTok{"purescript/}\SpecialCharTok{$\{}\NormalTok{name}\SpecialCharTok{\}}\StringTok{/src/**/*.purs"}\OperatorTok{;}
        \VariableTok{buildDir} \OperatorTok{=} \StringTok{"$HAKYLL\_DIR/\_purescript{-}build/}\SpecialCharTok{$\{}\NormalTok{name}\SpecialCharTok{\}}\StringTok{"}\OperatorTok{;}
        \VariableTok{mainModule} \OperatorTok{=} \StringTok{"}\SpecialCharTok{$\{}\NormalTok{buildDir}\SpecialCharTok{\}}\StringTok{/Main/index.js"}\OperatorTok{;}
        \VariableTok{outFile} \OperatorTok{=} \StringTok{"$HAKYLL\_DIR/\_purescript/}\SpecialCharTok{$\{}\NormalTok{name}\SpecialCharTok{\}}\StringTok{.js"}\OperatorTok{;}
      \KeywordTok{in}
      \StringTok{\textquotesingle{}\textquotesingle{}}
\StringTok{        mkdir {-}p }\SpecialCharTok{$\{}\NormalTok{buildDir}\SpecialCharTok{\}}
\StringTok{        purs compile }\SpecialCharTok{$\{}\BuiltInTok{toString}\NormalTok{ value.globs}\SpecialCharTok{\}}\StringTok{ }\SpecialCharTok{$\{}\NormalTok{srcGlob}\SpecialCharTok{\}}\StringTok{ {-}o }\SpecialCharTok{$\{}\NormalTok{buildDir}\SpecialCharTok{\}}
\StringTok{        chmod {-}R +w }\SpecialCharTok{$\{}\NormalTok{buildDir}\SpecialCharTok{\}}
\StringTok{        echo "import \{main\} from \textquotesingle{}}\SpecialCharTok{$\{}\NormalTok{mainModule}\SpecialCharTok{\}}\StringTok{\textquotesingle{}; main()" | esbuild {-}{-}bundle {-}{-}outfile=}\SpecialCharTok{$\{}\NormalTok{outFile}\SpecialCharTok{\}}\StringTok{ {-}{-}format=iife}
\StringTok{      \textquotesingle{}\textquotesingle{}}\OperatorTok{;}
  \KeywordTok{in}
\NormalTok{  pkgs.writeShellScriptBin}
    \StringTok{"rebuild{-}js"}
    \StringTok{\textquotesingle{}\textquotesingle{}}
\StringTok{      mkdir {-}p "$HAKYLL\_DIR/\_purescript";}
\StringTok{      }\SpecialCharTok{$\{}\NormalTok{lib.concatStringsSep }\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"} \OperatorTok{(}\NormalTok{lib.mapAttrsToList buildSingleDep inCode.purescript}\OperatorTok{)}\SpecialCharTok{\}}
\StringTok{    \textquotesingle{}\textquotesingle{}}\NormalTok{;}
\end{Highlighting}
\end{Shaded}

\texttt{writeShellScriptBin} is a convenient way to provide a derivation that
\emph{only} contains a single shell script in its \texttt{bin/} subdir. So if we
include it as a dependency in our development environment, this shell script
will be in scope. Keep in mind that the hakyll binary expects the output bundles
in the \texttt{\_purescript} directory.

In the script, \texttt{\$HAKYLL\_DIR} is our temporary working directory. We
iterate over each of the derivations in \texttt{inCode.purescript}, and for each
one, we create a temporary build directory and use \texttt{purs} to compile all
of the source files for the project (using the \texttt{globs} property that
\texttt{purifix} gives us, which contains globs of all the source files of the
dependencies as well, which are found in the global nix cache). And then we use
\texttt{esbuild} to manually generate the final bundle in the output directory.

And, to assemble the initial working directory, we have to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Copy all of the compiled purescript bundles into the temporary directory, from
  the derivation based on the current state at initialization time.
\item
  Simlink all of the source files (markdown, etc.) in the user's working
  directory to the temporary working directory, so any changes to the working
  directory will also be reflected in the temporary directory.
\item
  Provide all of the build tools through \texttt{nativeBuildInputs}.
\item
  Provide the user-accessible commands through \texttt{packages}.
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{devShells.default = pkgs.mkShell }\OperatorTok{\{}
  \VariableTok{shellHook} \OperatorTok{=} \StringTok{\textquotesingle{}\textquotesingle{}}
\StringTok{    export HAKYLL\_DIR=$(mktemp {-}d)}
\StringTok{    echo "Available commands: rebuild{-}js inCode{-}build"}
\StringTok{    echo "Hakyll working directory: \textbackslash{}$HAKYLL\_DIR"}

\StringTok{    mkdir {-}p $HAKYLL\_DIR/\_purescript}
\StringTok{    }\SpecialCharTok{$\{}\NormalTok{lib.concatStringsSep }\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}
    \OperatorTok{(}\NormalTok{lib.mapAttrsToList}
      \OperatorTok{(}\VariableTok{name}\OperatorTok{:} \VariableTok{value}\OperatorTok{:}
          \StringTok{\textquotesingle{}\textquotesingle{}}
\StringTok{          cp {-}a }\SpecialCharTok{$\{}\NormalTok{value.deps}\SpecialCharTok{\}}\StringTok{/output/. purescript/output}
\StringTok{          chmod {-}R +w purescript/output}
\StringTok{          cp }\SpecialCharTok{$\{}\NormalTok{value.bundle}\OperatorTok{{-}}\NormalTok{app}\SpecialCharTok{\}}\StringTok{ $HAKYLL\_DIR/\_purescript/}\SpecialCharTok{$\{}\NormalTok{name}\SpecialCharTok{\}}\StringTok{.js}
\StringTok{          \textquotesingle{}\textquotesingle{}}
      \OperatorTok{)}
\NormalTok{      inCode.purescript}
    \OperatorTok{)}\SpecialCharTok{\}}
\StringTok{    chmod {-}R +w $HAKYLL\_DIR/\_purescript}

\StringTok{    for srcDir in code{-}samples config copy css js latex scss static; do}
\StringTok{      ln {-}s "$PWD/$srcDir" $HAKYLL\_DIR}
\StringTok{    done}
\StringTok{  \textquotesingle{}\textquotesingle{}}\OperatorTok{;}
  \VariableTok{nativeBuildInputs} \OperatorTok{=} \OperatorTok{[}\NormalTok{ pkgs.esbuild pkgs.purescript }\OperatorTok{]}
    \OperatorTok{++}\NormalTok{ haskellFlake.devShell.nativeBuildInputs}
    \OperatorTok{++}\NormalTok{ lib.attrValues inCode.purescript}
    \OperatorTok{++} \BuiltInTok{map} \OperatorTok{(}\VariableTok{value}\OperatorTok{:}\NormalTok{ value.develop.buildInputs}\OperatorTok{)} \OperatorTok{(}\NormalTok{lib.attrValues inCode.purescript}\OperatorTok{);}
  \VariableTok{packages} \OperatorTok{=} \OperatorTok{[}
\NormalTok{    rebuild{-}js}
\NormalTok{    inCode.haskell}
  \OperatorTok{];}
\OperatorTok{\}}\NormalTok{;}
\end{Highlighting}
\end{Shaded}

\section{Conclusions}\label{conclusions}

Overall, I still feel like I'm hacking things together and figuring things out
as I go. So far, all I have done is automate a set-up that I have been doing
manually, in hopefully a way that is better for long-term maintainability.
However, hopefully as I learn more, I start to open up new doors and new ways of
thinking that enable to me to do things I haven't been able to do before.

Thank you for reading this --- whether you are looking to get started into nix,
or are an experienced veteran, I hope you were able to get something out of this
post. And, for those with experience, I would definitely appreciate any advice
concerning the things described in this post --- things I could have done
better, alternative approaches, or ways to be more idiomatic/maintainable.

\section{Signoff}\label{signoff}

Hi, thanks for reading! You can reach me via email at
\href{mailto:justin@jle.im}{\nolinkurl{justin@jle.im}}, or at twitter at
\href{https://twitter.com/mstk}{@mstk}! This post and all others are published
under the \href{https://creativecommons.org/licenses/by-nc-nd/3.0/}{CC-BY-NC-ND
3.0} license. Corrections and edits via pull request are welcome and encouraged
at \href{https://github.com/mstksg/inCode}{the source repository}.

If you feel inclined, or this post was particularly helpful for you, why not
consider \href{https://www.patreon.com/justinle/overview}{supporting me on
Patreon}, or a \href{bitcoin:3D7rmAYgbDnp4gp4rf22THsGt74fNucPDU}{BTC donation}?
:)

\end{document}
