\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{\textbf{#1}}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Introducing the mutable library},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
% Make links footnotes instead of hotlinks:
\renewcommand{\href}[2]{#2\footnote{\url{#1}}}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\title{Introducing the mutable library}

\begin{document}
\maketitle

\% Justin Le \% January 23, 2020

\emph{Originally posted on
\textbf{\href{https://blog.jle.im/entry/introducing-the-mutable-library.html}{in
Code}}.}

\textbf{mutable}: \href{https://mutable.jle.im/}{documentation} /
\href{http://hackage.haskell.org/package/mutable}{reference} /
\href{https://github.com/mstksg/mutable}{github}

(\emph{Note:} This post has been heavily revised to reflect
\emph{mutable-0.2.0.0}, as of July 2020. For reference,
\href{https://github.com/mstksg/inCode/blob/7c25dd3798955e8287d31774da6fe34015256b5a/entry/introducing-the-mutable-library.md}{the
original post} is available on github.)

I'm excited to announce the release of the
\emph{\href{https://mutable.jle.im/}{mutable}} library!

The library offers what I call \emph{beautiful mutable values}\footnote{Okay so
  I don't actually think the library is beautiful, I just like the way that
  ``beautiful mutable values'' sounds when you say it out loud.} --- automatic,
composable piecewise-mutable references for your data types. Sort of like an
automatically generated \texttt{MVector}, but for all your \texttt{ADT}s.

My high-level goal was a composable and overhead-free solution for dealing with
mutable values in Haskell in a type-safe and clean way. After all, why do
imperative languages have to have all the fun? In Haskell, we can have the best
of both worlds: efficient and clean mutable algorithms \emph{and} type safety.

The \href{https://mutable.jle.im/}{official documentation and homepage is here},
so it's a good read if you want to be introduced to how to use the library and
where it is most effective. But I'm going to use this blog post to talk about
\emph{why} I wrote the library, some of the neat things you can do with it, and
the techniques that went into writing it.

\section{Motivation}\label{motivation}

The original motivation for this comes from my development of
\emph{\href{https://backprop.jle.im/}{backprop}} and
\emph{\href{https://github.com/mstksg/backprop-learn}{backprop-learn}}, as I was
trying to adapt my
\href{https://blog.jle.im/entries/series/+functional-models.html}{Functional
Models} framework to efficient Haskell code.

To properly train Artificial Neural Networks with Haskell, you need to do a lot
of independent piecewise mutations to matrices and vectors. This becomes
inefficient, quickly, because you have to do a lot of copying in the process for
pure vectors and neural network weights. This problem also comes up for
efficient simulations that require mutating many different components
independently under a tight loop.

\subsection{Piecewise-Mutable}\label{piecewise-mutable}

First of all, what do I mean by ``piecewise-mutable''? Well, a simple example is
the mutable vector type, where piecewise-mutable edits are able to save a lot of
time and memory allocation.

If we want to edit the first item in a vector multiple times, this is extremely
inefficient with a pure vector:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{addFirst ::} \DataTypeTok{Vector} \DataTypeTok{Double} \OtherTok{{-}\textgreater{}} \DataTypeTok{Vector} \DataTypeTok{Double}
\NormalTok{addFirst xs }\OtherTok{=} \FunctionTok{iterate}\NormalTok{ incr xs }\OperatorTok{!!} \DecValTok{1000000}
  \KeywordTok{where}
\NormalTok{    incr v }\OtherTok{=}\NormalTok{ v }\OperatorTok{V.//}\NormalTok{ [(}\DecValTok{0}\NormalTok{, (v }\OperatorTok{V.!} \DecValTok{0}\NormalTok{) }\OperatorTok{+} \DecValTok{1}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

That's because \texttt{addFirst} will copy over the entire vector for every step
--- every single item, even if not modified, will be copied one million times.
It is \(O(n*l)\) in memory updates --- it is very bad for long vectors or large
matrices.

However, this is extremely efficient with a mutable vector:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{addFirst ::} \DataTypeTok{Vector} \DataTypeTok{Double} \OtherTok{{-}\textgreater{}} \DataTypeTok{Vector} \DataTypeTok{Double}
\NormalTok{addFirst xs }\OtherTok{=}\NormalTok{ runST }\OperatorTok{$} \KeywordTok{do}
\NormalTok{    v }\OtherTok{\textless{}{-}}\NormalTok{ V.thaw xs}
\NormalTok{    replicateM\_ }\DecValTok{1000000} \OperatorTok{$} \KeywordTok{do}
\NormalTok{        MV.modify v }\DecValTok{0}\NormalTok{ (}\OperatorTok{+} \DecValTok{1}\NormalTok{)}
\NormalTok{    V.freeze v}
\end{Highlighting}
\end{Shaded}

(this action is run in \texttt{ST}, the monad for mutable actions that is
provided by GHC)

This is because all of the other items in the vector are kept the same and not
copied-over over the course of one million updates. It is \(O(n+l)\) in memory
updates. It is very good even for long vectors or large matrices.

This situation is somewhat contrived, but it isolates a problem that many
programs face. A more common situation might be that you have two functions that
each modify different items in a vector in sequence, and you want to run them
many times interleaved, or one after the other.

\subsection{Composite Datatype}\label{composite-datatype}

That was an example of using piecewise mutability for vectors, but it's not
exactly scalable. That's because it always requires having a separate type for
the \emph{pure} type and the \emph{value} type. We're lucky enough to have one
for \texttt{Vector}\ldots but what about for our own custom types? That's a lot
of headache.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{TwoVec} \OtherTok{=} \DataTypeTok{TV}\NormalTok{ \{}\OtherTok{ tv1 ::} \DataTypeTok{Vector} \DataTypeTok{Double}
\NormalTok{                 ,}\OtherTok{ tv2 ::} \DataTypeTok{Vector} \DataTypeTok{Double}
\NormalTok{                 \}}
  \KeywordTok{deriving} \DataTypeTok{Generic}
\end{Highlighting}
\end{Shaded}

To use this in a ``piecewise-mutable'' way, we would need a separate ``mutable''
version:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{TwoVecRef}\NormalTok{ s }\OtherTok{=} \DataTypeTok{TVR}\NormalTok{ \{}\OtherTok{ tvr1 ::} \DataTypeTok{MVector}\NormalTok{ s }\DataTypeTok{Double}
\NormalTok{                       ,}\OtherTok{ tvr2 ::} \DataTypeTok{MVector}\NormalTok{ s }\DataTypeTok{Double}
\NormalTok{                       \}}
\end{Highlighting}
\end{Shaded}

Then we can do things like ``mutate only the first item in the first vector'' a
million times, and be efficient with it.

We'd have to write functions to ``thaw'' and ``freeze''

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{thawTwoVec ::} \DataTypeTok{TwoVec} \OtherTok{{-}\textgreater{}} \DataTypeTok{ST}\NormalTok{ s (}\DataTypeTok{TwoVecRef}\NormalTok{ s)}
\NormalTok{thawTwoVec (}\DataTypeTok{TV}\NormalTok{ x y) }\OtherTok{=} \DataTypeTok{TVR} \OperatorTok{\textless{}$\textgreater{}}\NormalTok{ V.thaw x }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ V.thaw y}

\OtherTok{freezeTwoVec ::} \DataTypeTok{TwoVecRef}\NormalTok{ s }\OtherTok{{-}\textgreater{}} \DataTypeTok{ST}\NormalTok{ s }\DataTypeTok{TwoVec}
\NormalTok{freezeTwoVec (}\DataTypeTok{TVR}\NormalTok{ u v) }\OtherTok{=} \DataTypeTok{TV} \OperatorTok{\textless{}$\textgreater{}}\NormalTok{ V.freeze u }\OperatorTok{\textless{}*\textgreater{}}\NormalTok{ V.freze v}
\end{Highlighting}
\end{Shaded}

It just doesn't scale in a composable way. You'd have to create a second version
of every data type.

\subsection{Solution}\label{solution}

The library provides the \texttt{Mutable} typeclass and the \texttt{GRef} type,
where \texttt{GRef\ s\ X} is the automatically derived piecewise-mutable version
of \texttt{X}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Mutable}\NormalTok{ s }\DataTypeTok{TwoVec} \KeywordTok{where}
    \KeywordTok{type} \DataTypeTok{Ref}\NormalTok{ s }\DataTypeTok{TwoVec} \OtherTok{=} \DataTypeTok{GRef}\NormalTok{ s }\DataTypeTok{TwoVec}
\end{Highlighting}
\end{Shaded}

The type \texttt{GRef\ s\ TwoVec} is \emph{exactly} the \texttt{TwoVecRef} that
we defined earlier: it is a tuple of two \texttt{MVector}s. It can do this
because \texttt{Vector} itself has a \texttt{Mutable} instance, where its
mutable version is \texttt{MVector}. \texttt{GRef\ s\ TwoVec} is essentially the
``MVector'' of \texttt{TwoVec}.

This now gives us
\texttt{thawRef\ ::\ TwoVec\ -\textgreater{}\ ST\ s\ (GRef\ s\ TwoVec)} and
\texttt{freezeRef\ ::\ GRef\ s\ TwoVec\ -\textgreater{}\ ST\ s\ TwoVec}, for
free, so we can write:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{addFirst ::} \DataTypeTok{TwoVec} \OtherTok{{-}\textgreater{}} \DataTypeTok{TwoVec}
\NormalTok{addFirst xs }\OtherTok{=}\NormalTok{ runST }\OperatorTok{$} \KeywordTok{do}
\NormalTok{    v }\OtherTok{\textless{}{-}}\NormalTok{ thawRef xs}
\NormalTok{    replicateM\_ }\DecValTok{1000000} \OperatorTok{$} \KeywordTok{do}
\NormalTok{      withField }\OperatorTok{\#}\NormalTok{tv1 v }\OperatorTok{$}\NormalTok{ \textbackslash{}u }\OtherTok{{-}\textgreater{}}
\NormalTok{        MV.modify u }\DecValTok{0}\NormalTok{ (}\OperatorTok{+} \DecValTok{1}\NormalTok{)}
\NormalTok{    freezeRef v}
\end{Highlighting}
\end{Shaded}

This will in-place edit only the first item in the \texttt{tv1} field one
million times, without ever needing to copy over the contents \texttt{tv2}.
Basically, it gives you a version of \texttt{TwoVec} that you can modify
in-place piecewise. You can compose two functions that each work piecewise on
\texttt{TwoVec}:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{mut1 ::} \DataTypeTok{Ref}\NormalTok{ s }\DataTypeTok{TwoVec} \OtherTok{{-}\textgreater{}} \DataTypeTok{ST}\NormalTok{ s ()}
\NormalTok{mut1 v }\OtherTok{=} \KeywordTok{do}
\NormalTok{    withField }\OperatorTok{\#}\NormalTok{tv1 v }\OperatorTok{$}\NormalTok{ \textbackslash{}u }\OtherTok{{-}\textgreater{}}
\NormalTok{      MV.modify u }\DecValTok{0}\NormalTok{ (}\OperatorTok{+} \DecValTok{1}\NormalTok{)}
\NormalTok{      MV.modify u }\DecValTok{1}\NormalTok{ (}\OperatorTok{+} \DecValTok{2}\NormalTok{)}
\NormalTok{    withField }\OperatorTok{\#}\NormalTok{tv2 v }\OperatorTok{$}\NormalTok{ \textbackslash{}u }\OtherTok{{-}\textgreater{}}
\NormalTok{      MV.modify u }\DecValTok{2}\NormalTok{ (}\OperatorTok{+} \DecValTok{3}\NormalTok{)}
\NormalTok{      MV.modify u }\DecValTok{3}\NormalTok{ (}\OperatorTok{+} \DecValTok{4}\NormalTok{)}

\OtherTok{mut2 ::} \DataTypeTok{Ref}\NormalTok{ s }\DataTypeTok{TwoVec} \OtherTok{{-}\textgreater{}} \DataTypeTok{ST}\NormalTok{ s ()}
\NormalTok{mut2 v }\OtherTok{=} \KeywordTok{do}
\NormalTok{    withField }\OperatorTok{\#}\NormalTok{tv1 v }\OperatorTok{$}\NormalTok{ \textbackslash{}u }\OtherTok{{-}\textgreater{}}
\NormalTok{      MV.modify u }\DecValTok{4}\NormalTok{ (}\OperatorTok{+} \DecValTok{1}\NormalTok{)}
\NormalTok{      MV.modify u }\DecValTok{5}\NormalTok{ (}\OperatorTok{+} \DecValTok{2}\NormalTok{)}
\NormalTok{    withField }\OperatorTok{\#}\NormalTok{tv2 v }\OperatorTok{$}\NormalTok{ \textbackslash{}u }\OtherTok{{-}\textgreater{}}
\NormalTok{      MV.modify u }\DecValTok{6}\NormalTok{ (}\OperatorTok{+} \DecValTok{3}\NormalTok{)}
\NormalTok{      MV.modify u }\DecValTok{7}\NormalTok{ (}\OperatorTok{+} \DecValTok{4}\NormalTok{)}

\OtherTok{doAMillion ::} \DataTypeTok{TwoVec} \OtherTok{{-}\textgreater{}} \DataTypeTok{TwoVec}
\NormalTok{doAMillion xs }\OtherTok{=}\NormalTok{ runST }\OperatorTok{$} \KeywordTok{do}
\NormalTok{    v }\OtherTok{\textless{}{-}}\NormalTok{ thawRef xs}
\NormalTok{    replicateM\_ }\DecValTok{1000000} \OperatorTok{$} \KeywordTok{do}
\NormalTok{      mut1 v}
\NormalTok{      mut2 v}
\NormalTok{    freezeRef v}
\end{Highlighting}
\end{Shaded}

The end result? You can now modify only a single component of your large
composite data type (and even single items in vectors in them) without making
nested copies every time.

\section{Neat Consequences}\label{neat-consequences}

\subsection{Mutable Sum Types}\label{mutable-sum-types}

While developing the library, I accidentally also stumbled into a way of
automatically deriving useful mutable sum types and data structures in Haskell.
This was more or less a complete accident --- I was writing the code to
automatically generate \texttt{GRef}, and needed to account for sum types
somehow. The result was actually useful!

For example, it is a publicly kept secret that Haskell's list type --- ``linked
lists'', are actually very different from the
\href{https://en.wikipedia.org/wiki/Linked_list}{mutable linked lists}
encountered as a standard data structure in languages like Java and C++. As it
turns out, using \texttt{GRef\ m\ {[}a{]}} gives us exactly the mutable linked
list type \ldots{} for free!

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{List}\NormalTok{ a }\OtherTok{=} \DataTypeTok{Nil} \OperatorTok{|} \DataTypeTok{Cons}\NormalTok{ a (}\DataTypeTok{List}\NormalTok{ a)}
  \KeywordTok{deriving}\NormalTok{ (}\DataTypeTok{Show}\NormalTok{, }\DataTypeTok{Generic}\NormalTok{)}
\KeywordTok{infixr} \DecValTok{5} \OtherTok{\textasciigrave{}Cons\textasciigrave{}}

\KeywordTok{instance} \DataTypeTok{Mutable}\NormalTok{ s a }\OtherTok{=\textgreater{}} \DataTypeTok{Mutable}\NormalTok{ m (}\DataTypeTok{List}\NormalTok{ a) }\KeywordTok{where}
    \KeywordTok{type} \DataTypeTok{Ref}\NormalTok{ s (}\DataTypeTok{List}\NormalTok{ a) }\OtherTok{=} \DataTypeTok{GRef}\NormalTok{ s (}\DataTypeTok{List}\NormalTok{ a)}
\end{Highlighting}
\end{Shaded}

Here we are re-implementing the \texttt{List} data structure from scratch just
to show that there is nothing arbitrary going on with the default list --- it
works for any appropriately defined ADT. We could even do binary trees!

Right away we can write functions to flesh out the API for a mutable linked
list. For example, a function to check if a linked list is empty:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} | Check if a mutable linked list is currently empty}
\NormalTok{isEmpty}
\OtherTok{    ::} \DataTypeTok{Mutable}\NormalTok{ s a}
    \OtherTok{=\textgreater{}} \DataTypeTok{Ref}\NormalTok{ s (}\DataTypeTok{List}\NormalTok{ a)}
    \OtherTok{{-}\textgreater{}} \DataTypeTok{ST}\NormalTok{ s }\DataTypeTok{Bool}
\NormalTok{isEmpty }\OtherTok{=}\NormalTok{ hasBranch (constrMB }\OperatorTok{\#}\NormalTok{\_Nil)}
\end{Highlighting}
\end{Shaded}

Here is a function to ``pop'' a mutable linked list, giving us the first value
and shifting the rest of the list up.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{popStack}
\OtherTok{    ::} \DataTypeTok{Mutable}\NormalTok{ s a}
    \OtherTok{=\textgreater{}} \DataTypeTok{Ref}\NormalTok{ s (}\DataTypeTok{List}\NormalTok{ a)}
    \OtherTok{{-}\textgreater{}} \DataTypeTok{ST}\NormalTok{ s (}\DataTypeTok{Maybe}\NormalTok{ a)}
\NormalTok{popStack xs }\OtherTok{=} \KeywordTok{do}
\NormalTok{    c }\OtherTok{\textless{}{-}}\NormalTok{ projectBranch (constrMB }\OperatorTok{\#}\NormalTok{\_Cons) xs}
\NormalTok{    forM c }\OperatorTok{$}\NormalTok{ \textbackslash{}(y, ys) }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      o }\OtherTok{\textless{}{-}}\NormalTok{ freezeRef y}
\NormalTok{      moveRef xs ys}
      \FunctionTok{pure}\NormalTok{ o}
\end{Highlighting}
\end{Shaded}

And a function to concatenate a second linked list to the end of a first one:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{concatLists}
\OtherTok{    ::} \DataTypeTok{Mutable}\NormalTok{ s a}
    \OtherTok{=\textgreater{}} \DataTypeTok{Ref}\NormalTok{ s (}\DataTypeTok{List}\NormalTok{ a)}
    \OtherTok{{-}\textgreater{}} \DataTypeTok{Ref}\NormalTok{ s (}\DataTypeTok{List}\NormalTok{ a)}
    \OtherTok{{-}\textgreater{}} \DataTypeTok{ST}\NormalTok{ s ()}
\NormalTok{concatLists l1 l2 }\OtherTok{=} \KeywordTok{do}
\NormalTok{    c }\OtherTok{\textless{}{-}}\NormalTok{ projectBranch consBranch l1}
    \KeywordTok{case}\NormalTok{ c }\KeywordTok{of}
      \DataTypeTok{Nothing}      \OtherTok{{-}\textgreater{}}\NormalTok{ moveRef l1 l2}
      \DataTypeTok{Just}\NormalTok{ (\_, xs) }\OtherTok{{-}\textgreater{}}\NormalTok{ concatLists xs l2}
\end{Highlighting}
\end{Shaded}

\subsection{Higher-Kinded Data}\label{higher-kinded-data}

I'm rather enamoured by the
``\href{https://reasonablypolymorphic.com/blog/higher-kinded-data/}{higher-kinded
data}'' pattern made popular by Sandy Maguire. It essentially eliminates the
need for explicit getters and setters by making the data type \emph{itself} the
thing that offers what you want, and you can get at it by just pattern matching.

Because of this, if your data type is written in the ``higher-kinded data''
pattern, then \texttt{MyType\ f} doubles as both the pure type \emph{and} the
mutable type, just by choice of \texttt{f}. \texttt{MyTypeF\ Identity} would be
the pure version, and \texttt{MyTypeF\ (RefFor\ m)} would be the mutable
version.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{MyTypeF}\NormalTok{ f }\OtherTok{=} \DataTypeTok{MTF}
\NormalTok{    \{}\OtherTok{ mtfInt    ::} \DataTypeTok{HKD}\NormalTok{ f }\DataTypeTok{Int}
\NormalTok{    ,}\OtherTok{ mtfDouble ::} \DataTypeTok{HKD}\NormalTok{ f }\DataTypeTok{Double}
\NormalTok{    ,}\OtherTok{ mtfVec    ::} \DataTypeTok{HKD}\NormalTok{ f (}\DataTypeTok{V.Vector} \DataTypeTok{Double}\NormalTok{)}
\NormalTok{    \}}
  \KeywordTok{deriving} \DataTypeTok{Generic}

\KeywordTok{type} \DataTypeTok{MyType\textquotesingle{}} \OtherTok{=} \DataTypeTok{MyTypeF} \DataTypeTok{Identity}

\KeywordTok{instance} \DataTypeTok{Mutable}\NormalTok{ s }\DataTypeTok{MyType\textquotesingle{}} \KeywordTok{where}
    \KeywordTok{type} \DataTypeTok{Ref}\NormalTok{ s }\DataTypeTok{MyType\textquotesingle{}} \OtherTok{=} \DataTypeTok{MyTypeF}\NormalTok{ (}\DataTypeTok{RefFor}\NormalTok{ s)}
\end{Highlighting}
\end{Shaded}

We can directly use it like a normal data type:

\begin{Shaded}
\begin{Highlighting}[]
\DataTypeTok{MTF} \DecValTok{3} \FloatTok{4.5}\NormalTok{ (V.fromList [}\DecValTok{1}\OperatorTok{..}\DecValTok{100}\NormalTok{])}
\OtherTok{    ::} \DataTypeTok{MyType\textquotesingle{}}
\end{Highlighting}
\end{Shaded}

But now, \texttt{MyTypeF\ (RefFor\ s)} literally has mutable references as its
fields. You can pattern match to get \texttt{rI\ ::\ MutVar\ s\ Int},
\texttt{rD\ ::\ MutVar\ s\ Double}, and \texttt{rV\ ::\ MVector\ s\ Double}

\begin{Shaded}
\begin{Highlighting}[]
\DataTypeTok{MTF}\NormalTok{ rI rD}\OtherTok{ rV ::} \DataTypeTok{MyTypeF}\NormalTok{ (}\DataTypeTok{RefFor}\NormalTok{ s)}
\end{Highlighting}
\end{Shaded}

and the accessors work as well:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{mtfVec}
\OtherTok{    ::} \DataTypeTok{MyTypeF}\NormalTok{ (}\DataTypeTok{RefFor}\NormalTok{ s)}
    \OtherTok{{-}\textgreater{}} \DataTypeTok{MVector}\NormalTok{ s }\DataTypeTok{Double}
\end{Highlighting}
\end{Shaded}

You can use it like:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{runST }\OperatorTok{$} \KeywordTok{do}
\NormalTok{    r}\OperatorTok{@}\NormalTok{(}\DataTypeTok{MTF}\NormalTok{ rI rD rV) }\OtherTok{\textless{}{-}}\NormalTok{ thawRef }\OperatorTok{$} \DataTypeTok{MTF} \DecValTok{0} \FloatTok{19.3}\NormalTok{ (V.fromList [}\DecValTok{1}\OperatorTok{..}\DecValTok{10}\NormalTok{])}

\NormalTok{    replicateM\_ }\DecValTok{1000} \OperatorTok{$} \KeywordTok{do}

        \CommentTok{{-}{-} rI is just the \textquotesingle{}Int\textquotesingle{} ref}
\NormalTok{        modifyMutVar rI (}\OperatorTok{+} \DecValTok{1}\NormalTok{)}

        \CommentTok{{-}{-} rV is the \textquotesingle{}MVector\textquotesingle{}}
\NormalTok{        MV.modify rV (}\OperatorTok{+}\DecValTok{1}\NormalTok{) }\DecValTok{0}

\NormalTok{    freezeRef r}

\CommentTok{{-}{-} =\textgreater{} MTF 1000 19.3 [1001.0, 2.0, 3.0, 4.0, 5.0, 6.0, 7.0, 8.0, 9.0, 10.0]}
\end{Highlighting}
\end{Shaded}

The ``mutable version'' of a type literally \emph{is} the ADT, if you use the
higher-kinded data pattern!

\subsection{A Polymorphic Picture}\label{a-polymorphic-picture}

One important thing to note when looking at the actual library --- the examples
in this post show the provided actions in \texttt{ST}, the mutable actions monad
provided by GHC. However, the library provides these actions polymorphic for all
\texttt{PrimMonad\ m}, an abstraction provided by the
\emph{\href{https://hackage.haskell.org/package/primitive}{primitive}} library
to generalize for all ``mutable monads'' (like \texttt{IO} and monad
transformers applied to \texttt{IO} and \texttt{ST}), as long as
\texttt{PrimState\ m\ \textasciitilde{}\ s}, so you can run them in whatever
useful mutable monads you'd like.

\section{Reflections on Generic}\label{reflections-on-generic}

This library is pretty much powered 95\% by GHC Generics, as the name
\texttt{GRef} implies. GHC Generics is probably one of the single most powerful
tools we have in Hasekll-the-language for writing typesafe abstractions and
eliminating all the boilerplate.

The structure of the \texttt{GRef} data type is completely determined by using
the \emph{GHC.Generics} \texttt{Rep} of an algebraic data type with a
\texttt{Generic} instance. It breaks apart the products and sums and turns them
into the mutable references you \emph{would} normally write by hand.

Writing \texttt{GRef} itself was actually very pleasant: it just involves
matching up generic pieces with the references they represent. ``What is the
reference for a constant value? What is the reference for a product type? What
is the reference for a sum type?'' And, in the process of answering those
questions, I ended up discovering something new (as shown in the section above
about mutable linked lists).

Generics also powers the \emph{higher-kinded data} based systems, which can add
a lot of syntactic niceness to everything if you decide to use it.

Still, I understand not everyone wants to restructure their data types in terms
of higher-kinded data \ldots{} there are a lot of practical issues to doing so,
and it doesn't really work well with nested data types. For that, I turned to
\emph{\href{https://hackage.haskell.org/package/generic-lens}{generic-lens}}.

\emph{generic-lens} is what powers the OverloadedLabels-based field accessor
methods that let you work with \texttt{GRef}s in a seamless way, by being able
to do \texttt{withField\ \#blah}, etc., instead of having to directly match on
the \texttt{GRef} value's internal contents (which can be messy, admittedly). It
also allows you to do \texttt{withPos\ @2} to get the second item in your
\texttt{GRef}, and \texttt{withTuple} to allow you to get the mutable fields in
your data type as a tuple.

I was originally going to implement the field accessors myself, looking to
\emph{generic-lens} for inspiration. However, when I looked at the library's
internals, I realized there was a lot more going on than I had originally
thought. But, looking at what was exported, I realized that the library was
well-designed enough that I could actually directly use its generic
implementations for \emph{mutable}! As a result, the field/position/tuple
accessor code actually required no mucking around with generics at all --- I
could leverage \emph{generic-lens}, which was powerful enough to allow me to
eliminate all of my generics code.

I strongly recommend anyone looking to do things involving generic access to
fields to look at \emph{generic-lens} to see if it can eliminate all your
generics code as well!

Unfortunately, I wasn't able to re-use the code for the ``constructor'' access
(as seen with \texttt{constrMB\ \#\_Cons} earlier) --- but I could use it as
inspiration to write my own. The library offers a very clean and well-written
pattern to doing things like this that I probably would have spent a long time
trying to figure out, if I had to do it from scratch.

\section{Next Steps}\label{next-steps}

I learned a lot from GHC Generics writing this library --- in a sense, the
library is pretty much completely an application of GHC Generics, without much
new concepts beyond that.

My next step is to equip \emph{backprop} to use \texttt{Mutable} instead of its
\texttt{Backprop} typeclass, so it can do in-place mutation of composite data
types for much faster backpropagation.

However, my newly gained experience with generics from writing this library can
actually do a lot to improve the ergonomics of \emph{backprop} as well --- in
particular, with \texttt{BVar}, which has always been very annoying to work
with, even with the lens-based API offered. Working with a \texttt{BVar} as if
it were a normal value has always been annoying, especially with product types.
There are a lot of ways GHC generics can help this, that I am now only learning
about. Check back soon --- hopefully I'll have something to show by then.

Until then, happy mutating! And please let me know if you find any interesting
applications of the library :D

\section{Signoff}\label{signoff}

Hi, thanks for reading! You can reach me via email at
\href{mailto:justin@jle.im}{\nolinkurl{justin@jle.im}}, or at twitter at
\href{https://twitter.com/mstk}{@mstk}! This post and all others are published
under the \href{https://creativecommons.org/licenses/by-nc-nd/3.0/}{CC-BY-NC-ND
3.0} license. Corrections and edits via pull request are welcome and encouraged
at \href{https://github.com/mstksg/inCode}{the source repository}.

If you feel inclined, or this post was particularly helpful for you, why not
consider \href{https://www.patreon.com/justinle/overview}{supporting me on
Patreon}, or a \href{bitcoin:3D7rmAYgbDnp4gp4rf22THsGt74fNucPDU}{BTC donation}?
:)

\end{document}
